---
title: "__Topic 11 - ggplot2: Grammar of Graphics__"
author: "Esam Mahdi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    self_contained: true
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

# Introduction to R Graphics

**Why use ggplot2?**

- **Consistent grammar** - Unlike base **R** plots, **ggplot2** uses a unified "grammar of graphics" approach

- **Layered construction** - Build plots step by step, making complex visualizations manageable

- **Beautiful defaults** - Produces publication-ready graphics with minimal code

- **Easy customization** - Modify almost every aspect of your plot

## The Grammar of Graphics

`ggplot2` is built on a simple idea: **every plot has three components**:

- **Data** - The dataset you want to visualize

- **Mapping (aes)** - How variables map to visual properties (x position, y position, color, size, etc.)

- **Geoms** - The geometric objects that represent your data (`points`, `bars`, `lines`, etc.)
  
```
ggplot(data = DATAFRAME, 
       mapping = aes(x = VAR1, y = VAR2)) +
  geom_TYPE()
```

# The `mpg` Dataset

The `mpg` dataset contains fuel economy data from 1999 to 2008 for 234 popular car models:

**Variable Descriptions**


| Variable | Type | Description |
|:---|:---|:---|
| `manufacturer` | Categorical | Car manufacturer (15 levels) |
| `model` | Categorical | Car model |
| `displ` | Quantitative | Engine displacement in liters |
| `year` | Categorical | Year of manufacture (1999 or 2008) |
| `cyl` | Categorical/Ordinal | Number of cylinders (4, 5, 6, 8) |
| `trans` | Categorical | Transmission type |
| `drv` | Categorical | Drive type: f = front, r = rear, 4 = 4wd |
| `cty` | Quantitative | City miles per gallon |
| `hwy` | Quantitative | Highway miles per gallon |
| `fl` | Categorical | Fuel type |
| `class` | Categorical | Vehicle class (7 levels) |

```
# Load required packages
library(tidyverse) # Recall tidyverse includes ggplot2 and dplyr

# Load the mpg dataset (comes with ggplot2)
data(mpg)

# Explore the dataset
glimpse(mpg)
```

```{r, echo = FALSE}
library(tidyverse) 
data(mpg)
```

**Step 1: The `ggplot()` Function**

The `ggplot()` function initializes the plot. Notice it creates an empty canvas:

```{r, fig.width=9, fig.height=5}
# Empty canvas - nothing appears yet!
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))
```

**Step 2: Add Geoms (Geometric Objects)**

Geoms are the actual visual elements. Add them with `+:`

```{r, fig.width=9, fig.height=5}
# Add points to create a scatterplot
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point()
```

**Interpretation:** There is a negative relationship between engine displacement and highway MPG. Cars with larger engines tend to have lower fuel efficiency.

```{r, fig.width=9, fig.height=5}
# Alternatively, you can get the same plot using any of the following methods:
# ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy))

# or 
# ggplot() + geom_point(data = mpg, mapping = aes(x = displ, y = hwy))

# or
# mpg %>% ggplot(mapping = aes(x = displ, y = hwy)) + geom_point()

# or
# mpg %>% ggplot() + geom_point(mapping = aes(x = displ, y = hwy))
```

# Geom Functions

| Function | Layers | Options |
|:---|:---|:---------|
| `geom_point()` | Scatterplot | `color (or colour)`, `alpha`, `shape`, `size` |
| `geom_line()` | Line graph | `color (or colour)`, `alpha`, `linetype`, `size` |
| `geom_jitter()` | Jittered points | `color (or colour)`, `size`, `alpha`, `shape` |
| `geom_bar()` | Bar chart | `color (or colour)`, `fill`, `alpha` |
| `geom_boxplot()` | Box plot | `color (or colour)`, `fill`, `alpha`, `notch`, `width` |
| `geom_histogram()` | Histogram | `color (or colour)`, `fill`, `alpha`, `linetype`, `binwidth` |
| `geom_smooth()` | Fitted line | `method`, `formula`, `color (or colour)`, `fill`, `linetype`, `size` |
| `geom_density()` | Density plot | `color (or colour)`, `fill`, `alpha`, `linetype` |
| `geom_hline()` | Horizontal lines | `color (or colour)`, `alpha`, `linetype`, `size` |
| `geom_vline()` | Vertical lines | `color (or colour)`, `alpha`, `linetype`, `size` |
| `geom_rug()` | Rug plot | `color (or colour)`, `side` |
| `geom_violin()` | Violin plot | `color (or colour)`, `fill`, `alpha`, `linetype` |
| `geom_text()` | Text annotations | See the help for this function |

**Step 3: Customize Appearance**

Modify colors, sizes, transparency, and theme:

```{r, fig.width=9, fig.height=5}
# Customized scatterplot
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(color = "#66C2A5", size = 4, alpha = 0.7, shape = 18) +
  theme_minimal() # Change the default gray theme to minimal theme
```

## Key Parameters For `geom_point():`

| Parameter | Description | Example |
|:---|:------------|:--------|
| `color (or colour)` | Changes the visual color of the object (points, lines, etc.) | `color = "steelblue"` (Default is black) |
| `size` | Controls the thickness of lines or the diameter of points | `size = 4` (Default is 1) |
| `alpha` | Sets the transparency level between 0 and 1 (0 is invisible, 1 is solid) | `alpha = 0.7` (Default is 1) |
| `shape` | Changes the style of the point. It takes integer values from 0 to 25 (e.g., circles, squares, triangles) | `shape = 18` (Default: `16` = filled circle) |

**Step 4: Add a Trend Line**

Use `geom_smooth()` to add lines of best fit:

```{r, fig.width=9, fig.height=5}
# Scatterplot with regression line (not I use %>%)
# se = FALSE to skip plotting the confidence interval of standard error and only draw the smoothed line
mpg %>%
  ggplot(mapping = aes(x = displ, y = hwy)) +
  geom_point(size = 4, alpha = 0.7) +
  geom_smooth(method = "lm", color = "darkred", se = FALSE) +
  theme_bw() # Black white theme
```

**Step 5: Map Variables to Aesthetics**

Instead of fixed colors, map variables to visual properties. Let's map `class` (vehicle class type) to color, and plot the best-fit line (linear regression) to each class data points:

```{r, fig.width=9, fig.height=5}
# Color points by vehicle class type
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point(size = 4, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic() # Classical theme
```

Now let's map `cyl` (cylinders) to color and shape, and plot the best of fit line to each number of cylinder data points:

```{r, fig.width=9, fig.height=5}
# Color points by number of cylinders and change the shapes
mpg %>%
ggplot(aes(x = displ, y = hwy, color = factor(cyl), shape = factor(cyl))) +
  geom_point(size = 4, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

**Step 6: Add Labels**

Make your plot informative with `labs()`:

```{r, fig.width=9, fig.height=5}
# Add informative labels and note that I map the class of the cars to color and size
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, colour = class, size = class)) +
  geom_point() +
  labs(
    title = "Relationship Between Engine Size and Highway MPG",
    subtitle = "By Vehicle Class",
    x = "Engine Displacement (Liters)",
    y = "Highway Miles Per Gallon",
    caption = "Source: EPA Fuel Economy Data (1999-2008)") +
  theme_light() # Light theme
```

**Interpretation:** As engine displacement increases, highway (miles per gallon) MPG generally decreases (negative correlation). Different vehicle classes cluster in different regions:

- **SUVs/pickups** → Large engines, low MPG

- **Compact/subcompact** → Small engines, high MPG

- **Midsize** → Middle ranges

**Step 7: Modify Scales**

Control axes with `scale_*()` functions:

```{r, fig.width=9, fig.height=5}
# Customize scales
p1 <- ggplot(data = mpg, mapping = aes(displ, hwy, color = class)) +
  geom_point(size = 4, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red", size = 0.8) +
  labs(
    title = "Engine Size vs Highway MPG",
    x = "Engine Displacement (Liters)",
    y = "Highway MPG"
  ) +
  scale_x_continuous(breaks = seq(1, 7, 1)) +
  scale_y_continuous(breaks = seq(10, 45, 5)) 

p1 + theme_dark() # Dark theme 
``` 

# Faceting: Multiple Plots by Group

One of ggplot2's most powerful features is **faceting** - creating multiple panels based on a categorical variable.

## Facet Wrap (Single Variable)

```{r, fig.width=9, fig.height=5}
# Facet by drive type
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "steelblue", size = 4, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_wrap(~drv, ncol = 3) +
  labs(
    title = "Engine Size vs Highway MPG by Drive Type",
    x = "Engine Displacement (L)",
    y = "Highway MPG") +
  theme_bw()
```  

**Interpretation:** The negative relationship between engine size and MPG is strongest for front-wheel drive cars.

**Parameters for faceting:**

| Function | Description | Example |
|:---|:---------|:---|
| `facet_wrap()` | Creates a 1D ribbon of panels for each level of a variable | `facet_wrap(~variable)` |
| `facet_grid()` | Creates a 2D grid of panels defined by two variables | `facet_grid(row ~ col)` |
| `ncol` | Sets the specific number of columns in the layout | `ncol = 3`|
| `nrow` | Sets the specific number of rows in the layout | `nrow = 2` |
| `scales` | Determines if axis scales are shared or independent across panels | `scales = "free"` |

## Facet Grid (Two Variables)

```{r, fig.width=9, fig.height=7}
# Facet by year and drive type
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(size = 4, alpha = 0.7, aes(color = class)) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_grid(year ~ drv, 
             labeller = labeller(drv = c("4" = "4WD", "f" = "Front", "r" = "Rear"))) +
  labs(
    title = "Engine Size vs Highway MPG",
    subtitle = "Faceted by Year and Drive Type",
    x = "Engine Displacement (L)",
    y = "Highway MPG") +
  theme_bw() +
  theme(legend.position = "bottom")
``` 

**Scales argument in `facet_wrap()` and `facet_grid()`**

The `scales` argument in `facet_wrap()` and `facet_grid()` controls whether axis scales are shared ("fixed", default) or allowed to vary (`"free"`, `"free_x"`, `"free_y"`) across panels, enabling easier comparison within or between subsets.
This is especially useful for comparing plots with widely different data ranges

```{r, fig.width=9, fig.height=7}
# Free x - each panel has its own x-axis scale
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(size = 4, alpha = 0.7, aes(color = class)) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_grid(year ~ drv, 
             scales = "free_x",  # Free x-axis only
             labeller = labeller(drv = c("4" = "4WD", "f" = "Front", "r" = "Rear"))) +
  labs(
    title = "Engine Size vs Highway MPG (Free X-axis)",
    subtitle = "Each panel has its own x-axis scale",
    x = "Engine Displacement (L)",
    y = "Highway MPG") +
  theme_bw() +
  theme(legend.position = "bottom")
``` 

**Scales options in facetting**

| `scales` argument | Description | When to Use |
|:---|:---|:---|
| `"fixed"` (default) | Every panel uses the exact same x and y scale | When comparing panels directly and data ranges are similar |
| `"free_x"` | Each panel has its own x-axis scale; y-axis remains the same | When panels have different x-axis ranges but comparable y-values |
| `"free_y"` | Each panel has its own y-axis scale; x-axis remains the same | When panels have different y-axis ranges but comparable x-values |
| `"free"` | Both x and y scales are independent for each panel | When panels have very different ranges for both axes |

# Bar Charts with `geom_bar()`

The `geom_bar()` is designed to visualize the distribution of a single (or multiple) **categorical variable** by calculating and displaying the **total count** for each category as bar heights.

**Basic bar chart**

```{r, fig.width=9, fig.height=5}
# Bar chart for manufacturer (categorical variable) distribution
ggplot(mpg, aes(x = manufacturer)) +
  geom_bar() 
```

**Interpretation:** `Dodge`, `Toyota`, and `Volkswagen` have the most models in the dataset.

## Customized Bar Chart

```{r, fig.width=9, fig.height=5}
# Enhanced bar chart for vehicle class variable
ggplot(mpg, aes(x = class)) +
  geom_bar(fill = "steelblue", color = "black", width = 0.7) +
  labs(
    title = "Distribution of Vehicle Classes",
    subtitle = "MPG Dataset (1999-2008)",
    x = "Vehicle Class",
    y = "Count") +
  theme_bw() 
```

**Interpretation:** `SUV` and `compact` cars are most common, while `2seaters` are rarest.

## Key Parameters For `geom_bar()`:

| Parameter | Description | Example |
|:---|:------|:---|
| `fill` | Changes the interior color of the bar | `fill = "steelblue"` |
| `color` | Changes the color of the bar's border/outline | `color = "black"` |
| `width` | Controls the thickness of the bars (relative to the x-axis) | `width = 0.7` |
| `alpha` | Sets the transparency (0 is invisible, 1 is solid) | `alpha = 0.8` |
| `stat` | ???? | `stat = "identity"` |

## Bar Chart with Percentages

```{r, fig.width=9, fig.height=5}
# Percentages on y-axis for drive type variable
# Load the scales package first
library(scales)

# Now the code will work
ggplot(mpg, aes(x = drv, y = after_stat(count / sum(count)))) +
    geom_bar(fill = "tomato", color = "black") +
    scale_y_continuous(labels = percent) +
    labs(
        title = "Drive Type Distribution",
        x = "Drive Type (f = front, r = rear, 4 = 4wd)",
        y = "Percentage"
    ) +
    theme_bw()
```

**Interpretation:** `Front-wheel` drive is the most common, followed by `4WD`.

## Sorting Bars by Frequency

```{r, fig.width=9, fig.height=5}
# Sort bars in descending order for cylinders variable 
mpg %>%
  count(cyl) %>%
  mutate(cyl = reorder(factor(cyl), -n)) %>%
  ggplot(aes(x = cyl, y = n)) +
  geom_bar(stat = "identity", fill = "cornflowerblue") +
  geom_text(aes(label = n), vjust = -0.5, size = 5) +
  labs(x = "Number of Cylinders", y = "Count", title = "Engine Cylinder Distribution (Sorted)") +
  theme_bw() 
```

**Interpretation:** `4-cylinder` and `6-cylinder` engines are most common. `5-cylinder` engines are rare.

## Horizontal Bar Chart

```{r, fig.width=9, fig.height=5}
# Horizontal bar chart for manufacturers variable
manufacturer_counts <- mpg %>% 
  count(manufacturer) %>%
  arrange(n)

ggplot(manufacturer_counts, aes(x = reorder(manufacturer, n), y = n)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  coord_flip() +
  labs(x = "", y = "Number of Models", title = "Car Models by Manufacturer") +
  theme_bw() 
```

**Interpretation:** `Dodge`, `Toyota`, and `Volkswagen` have the most models. Mercury and Porsche have the fewest.

# Histograms with `geom_histogram()`

The `geom_histogram()` function is designed to visualize the distribution of a single (or more) **continuous quantitative variable** by splitting the data into equal-width bins and plotting the **counts of observations** within each bin using **bars**.

**Basic histogram**

```{r, fig.width=9, fig.height=5}
# Basic histogram 
ggplot(mpg, aes(x = hwy)) +
  geom_histogram() 
```

**Interpretation:** Most cars get between 20-30 MPG on the highway.

**Customized histogram - city MPG**

```{r, fig.width=9, fig.height=5}
# Customized histogram for city MPG
ggplot(mpg, aes(x = cty)) +
  geom_histogram(
    fill = "lightblue",
    color = "white",
    bins = 20,
    alpha = 0.8
  ) +
  labs(
    title = "Distribution of City Fuel Economy",
    x = "City Miles Per Gallon",
    y = "Frequency") +
  theme_bw() 
```

## Key Parameters For `geom_histogram()`:

| Parameter | Description | Example |
|:---|:---|:---|
| `bins` | Sets the total number of bars (bins) in the plot | `bins = 20` |
| `binwidth` | Sets the width of each bar in the units of the x-axis variable | `binwidth = 5` |
| `fill` | Changes the interior color of the bars | `fill = "lightblue"` |
| `color` | Changes the color of the border/outline between bars | `color = "white"` |
| `alpha` | Sets the transparency level (0 to 1) | `alpha = 0.8` |

## Controlling Bin Width

```{r, fig.width=9, fig.height=5}
# Compare different bin widths (for engine displacement variable)
p1 <- ggplot(mpg, aes(x = displ)) +
  geom_histogram(fill = "skyblue", bins = 5) +
  labs(title = "5 bins (too few)", x = "Engine Displacement (L)", y = "Count") +
  theme_minimal()

p2 <- ggplot(mpg, aes(x = displ)) +
  geom_histogram(fill = "skyblue", bins = 15) +
  labs(title = "15 bins (good)", x = "Engine Displacement (L)", y = "Count") +
  theme_minimal()

p3 <- ggplot(mpg, aes(x = displ)) +
  geom_histogram(fill = "skyblue", bins = 40) +
  labs(title = "40 bins (too many)", x = "Engine Displacement (L)", y = "Count") +
  theme_minimal()

library(patchwork) # To combine separate ggplot2 plots into a single aligned multi-panel figure 
p1 + p2 + p3
```

**Interpretation:** Engine displacement is bimodal, with peaks around 2.0L and 5.0-6.0L.

## Histogram with Density Curve

```{r, fig.width=9, fig.height=5}
# Histogram with density overlay (for highway MPG by year variable)
ggplot(mpg, aes(x = hwy, y = after_stat(density), fill = factor(year))) +
  geom_histogram(color = "green", bins = 20, alpha = 0.6, position = "identity") +
  geom_density(size = 1.2, alpha = 0.3) +
  scale_fill_manual(values = c("steelblue", "coral"), name = "Year") +
  labs(
    title = "Highway MPG Distribution by Year",
    x = "Highway MPG",
    y = "Density") +
  theme_bw() 
```
  
**Interpretation:** 2008 models show improved fuel economy compared to 1999 models.

## Adding Summary Statistics

```{r, fig.width=9, fig.height=5}
# Calculate statistics (for engine displacement variable)
mean_displ <- mean(mpg$displ)
median_displ <- median(mpg$displ)

# Plot the histogram with mean and median lines (for engine displacement variable)
ggplot(mpg, aes(x = displ)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 15, alpha = 0.7) +
  geom_vline(xintercept = mean_displ, color = "red", size = 1.2, linetype = "dashed") +
  geom_vline(xintercept = median_displ, color = "blue", size = 1.2, linetype = "dotted") +
  annotate("text", x = 4, y = 35, 
           label = paste("Mean =", round(mean_displ, 1), "L"), color = "red") +
  annotate("text", x = 4, y = 30, 
           label = paste("Median =", round(median_displ, 1), "L"), color = "blue") +
  labs(
    title = "Engine Displacement Distribution",
    x = "Engine Displacement (Liters)",
    y = "Frequency") +
  theme_bw() 
```

## Histogram with Faceting

```{r, fig.width=9, fig.height=7}
ggplot(mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  facet_wrap(~ drv, ncol = 1, scales = "free") +
  labs(
    title = "Highway MPG Distribution by Drive Type",
    x = "Highway Miles Per Gallon",
    y = "Count"
  ) +
  theme_bw()
```

# Boxplots with `geom_boxplot()`

A **boxplot** (or box-and-whisker plot) is a standardized way of displaying the distribution of a variable based on a **five-number summary:** 

| Component | Description | Calculation / Definition |
|:---|:---|:---|
| **Minimum** | Smallest value | $Q1 - 1.5 \times IQR$ |
| **Q1** (First Quartile) | 25th percentile | 25% of data falls below this value |
| **Median** (Q2) | 50th percentile | The middle value of the dataset |
| **Q3** (Third Quartile) | 75th percentile | 75% of data falls below this value |
| **Maximum** | Largest value | $Q3 + 1.5 \times IQR$ |
| **IQR** | Interquartile Range | $Q3 - Q1$ |

**The boxplot will display:**

- **Median** (middle line inside the box) - the 50th percentile

- **Box** spanning from Q1 (25th percentile) to Q3 (75th percentile) - the interquartile range (IQR)

- **Whiskers** extending to:

  - Q1 - 1.5×IQR (lower bound)

  - Q3 + 1.5×IQR (upper bound)

- **Points** beyond the whiskers (potential **outliers**)

**Basic Boxplot**

```{r, fig.width=9, fig.height=5}
# Basic single boxplot - highway MPG distribution
ggplot(mpg, aes(x = "", y = hwy)) + # Note: "ggplot(mpg, aes(hwy))" produces a boxplot with the coordinates flipped horizontally
  geom_boxplot() 
```

We can conclude that the distribution (of **all cars**) is **left-skewed** (because the median is closer to the right side of the box) with two high-end potential **outliers**.  
    
## Grouped Boxplots (Categorical X-axis)

```{r, fig.width=9, fig.height=5}
# Boxplot by drive type (categorical variable)
ggplot(mpg, aes(x = drv, y = hwy, fill = drv)) +
  geom_boxplot() +
  scale_x_discrete(labels = c("4" = "4WD", "f" = "Front", "r" = "Rear")) +
  labs(
    title = "Highway MPG Distribution by Drive Type",
    x = "Drive Type",
    y = "Highway MPG",
    fill = "Drive Type") +
  theme_bw()
```

## Customizing Boxplots

### Changing Colors and Appearance

```{r, fig.width=9, fig.height=5}
# Customized boxplot with specific colors
ggplot(mpg, aes(x = drv, y = hwy, fill = drv)) +
  geom_boxplot(
    alpha = 0.7,
    color = "black",
    size = 0.8,
    outlier.color = "red",
    outlier.size = 4,
    outlier.shape = 16
  ) +
  scale_fill_manual(values = c("lightblue", "lightgreen", "coral"),
                    labels = c("4WD", "Front", "Rear")) +
  scale_x_discrete(labels = c("4" = "4WD", "f" = "Front", "r" = "Rear")) +
  labs(
    title = "Customized Boxplot: Highway MPG by Drive Type",
    x = "Drive Type",
    y = "Highway MPG",
    fill = "Drive Type") +
  theme_bw()
```

### Horizontal Boxplots

```{r, fig.width=9, fig.height=5}
# Horizontal orientation (useful when category labels are long)
ggplot(mpg, aes(x = class, y = hwy, fill = class)) +
  geom_boxplot(alpha = 0.7) +
  coord_flip() +  # Flip coordinates to make horizontal
  labs(
    title = "Highway MPG Distribution by Vehicle Class",
    x = "",
    y = "Highway MPG",
    fill = "Vehicle Class") +
  theme_bw() +
  theme(legend.position = "none")
```

### Boxplot with Faceting

```{r, fig.width=9, fig.height=7}
# Boxplot of highway MPG by vehicle class, faceted by drive type
mpg %>% 
  ggplot(aes(x = class, y = hwy, fill = class)) +
  geom_boxplot(outlier.size = 2) +
  facet_wrap(~drv, nrow = 3, scales = "free_x",
             labeller = labeller(drv = c("4" = "4WD", "f" = "Front", "r" = "Rear"))) +
  theme_bw() 
```

### Comprehensive Comparison: Boxplot vs Violin Plot

```{r, fig.width=9, fig.height=5}
# Compare boxplot with violin plot
library(patchwork)

# Boxplot
p_box <- ggplot(mpg, aes(x = drv, y = hwy, fill = drv)) +
  geom_boxplot(alpha = 0.7) +
  scale_x_discrete(labels = c("4" = "4WD", "f" = "Front", "r" = "Rear")) +
  labs(title = "Boxplot", x = "Drive Type", y = "Highway MPG") +
  theme_minimal() +
  theme(legend.position = "none")

# Violin plot (shows full distribution shape)
p_violin <- ggplot(mpg, aes(x = drv, y = hwy, fill = drv)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.2, alpha = 0.8) +
  scale_x_discrete(labels = c("4" = "4WD", "f" = "Front", "r" = "Rear")) +
  labs(title = "Violin Plot with Boxplot Overlay", x = "Drive Type", y = "Highway MPG") +
  theme_minimal() +
  theme(legend.position = "none")

# Arrange side by side
p_box + p_violin
```

**Note:** **Boxplots** provide a concise **summary of distribution** through quartiles and outliers, making them ideal for comparing multiple groups. **Violin plots** extend this by showing the full **probability density**, revealing multimodal distributions that boxplots might miss. Use **boxplots** for precise quantile comparisons; use **violin plots** when understanding the complete shape of the distribution matters.

## Key Parameters For `geom_boxplot()`

| Parameter | Description | Example |
|:---|:---|:---|
| `fill` | Changes the interior color of the box | `fill = "steelblue"` |
| `color` | Changes the color of the box and whisker borders | `color = "black"` |
| `alpha` | Sets the transparency level (0 to 1) | `alpha = 0.7` |
| `size` | Controls the thickness of the box and whisker lines | `size = 0.8` |
| `outlier.color` | Sets the specific color for outlier points | `outlier.color = "red"` |
| `outlier.size` | Sets the size of the outlier points | `outlier.size = 4` |
| `outlier.shape` | Sets the visual symbol/shape for outliers | `outlier.shape = 16` |
| `width` | Controls the width of the box relative to the x-axis | `width = 0.5` |
| `notch` | Adds a "waist" representing the median confidence interval | `notch = TRUE` |
| `varwidth` | Makes box width proportional to the sample size of the group | `varwidth = TRUE` |

## Stacked, Grouped, and Filled Bar Charts

Beyond simple bar charts, `geom_bar()` offers powerful ways to display relationships between **categorical variables** using the `fill` aesthetic and `position` argument.

### Stacked Bar Chart

A stacked bar chart shows the composition of each category by stacking segments on top of each other. The total height represents the overall count, while segment heights show the contribution of each subgroup.

```{r, fig.width=9, fig.height=5}
# Stacked bar chart - showing drive type composition within each vehicle class
ggplot(mpg, aes(x = class, fill = drv)) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("4" = "#66C2A5", "f" = "#FC8D62", "r" = "#8DA0CB"),
                    labels = c("4" = "4WD", "f" = "Front", "r" = "Rear")) +
  labs(
    title = "Stacked Bar Chart: Drive Type Distribution by Vehicle Class",
    subtitle = "Each bar totals to 100% of vehicles in that class",
    x = "Vehicle Class",
    y = "Count",
    fill = "Drive Type") +
  theme_bw() 
```

**Note:** The stacked bar chart shows that **SUVs** and **compact** cars have the **highest total counts**. **Front-wheel** drive (**orange**) dominates all vehicle classes, especially **compact**, **midsize** and **subcompact** cars. **4WD** (**green**) appears almost exclusively in **SUVs** and **pickups**, while **rear-wheel** drive (**blue**) is most common in **2-seaters** and **pickups**. However, comparing non-bottom segments (like **4WD** counts) across different bars is difficult because they don't share a common baseline.

### Grouped (Dodged) Bar Chart

A grouped bar chart places bars for different subgroups side-by-side within each category, making it easy to compare subgroup values both within and across categories.

```{r, fig.width=9, fig.height=5}
# Grouped bar chart - comparing drive type counts across vehicle classes
ggplot(mpg, aes(x = class, fill = drv)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("4" = "#66C2A5", "f" = "#FC8D62", "r" = "#8DA0CB"),
                    labels = c("4" = "4WD", "f" = "Front", "r" = "Rear")) +
  labs(
    title = "Grouped Bar Chart: Drive Type Comparison by Vehicle Class",
    subtitle = "Side-by-side bars enable direct comparison of subgroups",
    x = "Vehicle Class",
    y = "Count",
    fill = "Drive Type") +
  theme_bw() 
```

**Note:** Grouped bar charts excel at facilitating direct comparisons between subgroups across categories because the side-by-side positioning puts related bars adjacent to each other. They work best when the number of subgroups is small (2-4) to avoid cluttered and unreadable visualizations.

### Filled (Proportional) Bar Chart

A filled bar chart normalizes each bar to the same total height (100%), turning the y-axis into percentages. This emphasizes the relative proportion of subgroups within each category rather than the absolute counts.

```{r, fig.width=9, fig.height=5}
# Filled bar chart - showing proportional composition of drive types within each vehicle class
library(scales)  # For percent formatting

ggplot(mpg, aes(x = class, fill = drv)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("4" = "#66C2A5", "f" = "#FC8D62", "r" = "#8DA0CB"),
                    labels = c("4" = "4WD", "f" = "Front", "r" = "Rear")) +
  labs(
    title = "Filled Bar Chart: Proportion of Drive Types by Vehicle Class",
    subtitle = "Each bar represents 100% of vehicles in that class, showing composition only",
    x = "Vehicle Class",
    y = "Percentage",
    fill = "Drive Type") +
  theme_bw() 
```

**Note:** Filled bar charts (using `position = "fill"`) are ideal when comparing proportional composition across categories while ignoring differences in category sizes. They answer questions like **"Which vehicle class has the highest proportion of 4WD vehicles?"** rather than **"Which class has the most 4WD vehicles?"**

## Position Arguments Summary

| Position | Description | Best Used When |
|:---|:---|:---|
| `"stack"` | Bars are stacked on top of each other | Comparing totals and subgroup contributions simultaneously |
| `"dodge"` | Bars are placed side-by-side | Making direct comparisons between subgroup values within and across categories |
| `"fill"` | Bars are normalized to the same height (100%) | Comparing proportional composition across categories regardless of total count differences |
| `"identity"` | Bars overlap (use with transparency) | Visualizing overlapping distributions (less common for bar charts) |

# Key Summaries

## The Grammar of Graphics in a Nutshell

Every **ggplot2** plot follows this template:

```
ggplot(data = DATAFRAME, aes(x = VAR1, y = VAR2)) +
    geom_TYPE() +
  customization_functions()
```

## Choosing the Right Plot Type

| Goal | Plot Type | Geometry Function |
|:---|:---|:---|
| Show relationship between two continuous variables | Scatterplot | `geom_point()` |
| Show trend/relationship | Trend line | `geom_smooth()` |
| Distribution of a continuous variable | Histogram | `geom_histogram()` |
| Distribution of a categorical variable | Bar chart | `geom_bar()` |
| Compare distributions across groups | Boxplot or Violin plot | `geom_boxplot()` or `geom_violin()` |
| Show composition of categories | Stacked/dodged/filled bar chart | `geom_bar(position = ...)` |
| Create multiple small plots | Faceting | `facet_wrap()` or `facet_grid()` |

## Essential ggplot2 Functions

| Function | Purpose | Example with `mpg` |
|:---|:---|:---|
| `ggplot()` | Initialize the plot | `ggplot(mpg, aes(x = displ, y = hwy))` |
| `geom_point()` | Create a scatterplot | `geom_point(color = "blue")` |
| `geom_bar()` | Create a bar chart| `geom_bar(fill = "steelblue")` |
| `geom_histogram()` | Create a histogram| `geom_histogram(bins = 20)` |
| `geom_density()` | Create a density plot| `geom_density(fill = "coral")` |
| `geom_boxplot` | Create a box-plot| `geom_boxplot(y = hwy)` |
| `geom_smooth()` | Add a trend line or line of best fit| `geom_smooth(method = "lm")`|
| `facet_wrap()` | Split into multiple panels by a variable| `facet_wrap(~drv)`|
| `labs()` | Add titles and labels| `labs(title = "My Title")`|
| `theme_*()` | Apply a visual theme| `theme_minimal()`|

## Aesthetic Mapping vs. Setting

| Concept | Syntax | Description | Example |
|:---|:---|:------|:---|
| **Mapping** | Inside `aes()` | Data variable determines visual property | `aes(color = class)` |
| **Setting** | Outside `aes()` | You manually assign a fixed value | `color = "steelblue"` |

### Common Aesthetic Mappings (Inside `aes()`)

| Aesthetic | Maps to... | Example with `mpg` |
|:---|:---|:---|
| `x` | X-axis position | `aes(x = displ)` |
| `y` | Y-axis position | `aes(y = hwy)` |
| `color` | Color of points or outlines | `aes(color = class)` |
| `fill` | Interior color of shapes (bars, boxes) | `aes(fill = drv)` |
| `size` | Size of points or lines| `aes(size = cyl)`|
| `shape` | The visual symbol/shape of a point| `aes(shape = year)` |
| `alpha` | Transparency level based on data values| `aes(alpha = cty)`|

## Position Adjustments for Bar Charts

| Position | Code | What It Does |
|:---|:---|:---|
| Stack | `position = "stack"` (default) | Bars stacked on top of each other |
| Dodge | `position = "dodge"` | Bars placed side-by-side |
| Fill | `position = "fill"` | Bars normalized to 100% height |

## Key Parameters Reference

**For `geom_point()`:** `color`, `size`, `alpha`, `shape`

**For `geom_bar()` / `geom_histogram()`:** `fill` (interior color), `color` (border color), `width`/`binwidth`

**For `geom_boxplot()`:** `fill`, `color`, `outlier.color`, `outlier.size`, `notch`

**For Faceting:**

- `facet_wrap(~variable, ncol = X)`
- `facet_grid(row_variable ~ col_variable)`
- `scales = "free"`, `"free_x"`, `"free_y"`

**For Labels:** `labs(title = "...", subtitle = "...", x = "...", y = "...", caption = "...")`

**For Themes:** `theme_bw()`, `theme_minimal()`, `theme_classic()`, `theme_light()`, `theme_dark()`

## Common Mistakes to Avoid

1. **Forgetting `aes()`:**  Mapping variables requires `aes()`, setting fixed aesthetics does not

2. **Using `=` instead of `==` in `aes()`:**  Use `==` for conditions outside `aes()` only

3. **Putting `+` at the beginning of a line:**  The `+` must end the previous line

4. **Not specifying `stat = "identity"`:**  When using pre-summarized data with `geom_bar()`

5. **Confusing `color` and `fill`:** `color` is for outlines/borders, `fill` is for interior

# Exercises (Using `iris` Dataset)

The `iris` dataset contains 150 observations of iris flowers with 5 variables:

| Variable | Type | Description |
|:---|:---|:------|
| `Sepal.Length` | Quantitative | Sepal length in centimeters |
| `Sepal.Width` |	Quantitative | Sepal width in centimeters |
| `Petal.Length` |	Quantitative | Petal length in centimeters |
| `Petal.Width` |	Quantitative | Petal width in centimeters |
| `Species` |	Categorical | Species of iris (**setosa**, **versicolor**, **virginica**) |

## Exercise 1: Basic Scatterplot with Customization

Using the `iris` dataset, create a scatterplot showing the relationship between **sepal length** (`Sepal.Length`) and **petal length** (`Petal.Length`). Customize the points with:

- Blue color

- Size 3

- 70% transparency (`alpha = 0.7`)

- Diamond shape (`shape = 18`)

Add appropriate axis labels and a title "Sepal Length vs Petal Length in Iris Flowers".

```
# Your code here:
```

## Solution

```{r, fig.width=9, fig.height=5}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point(color = "blue", size = 3, alpha = 0.7, shape = 18) +
  labs(
    title = "Sepal Length vs Petal Length in Iris Flowers",
    x = "Sepal Length (cm)",
    y = "Petal Length (cm)"
  ) 
```

## Exercise 2: Adding a Trend Line

Starting from Exercise 1, add a linear regression trend line (`method = "lm"`) in red without showing the confidence interval (`se = FALSE`).

```
# Your code here:
```

## Solution

```{r, fig.width=9, fig.height=5}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point(color = "blue", size = 3, alpha = 0.7, shape = 18) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Sepal Length vs Petal Length in Iris Flowers",
    x = "Sepal Length (cm)",
    y = "Petal Length (cm)"
  ) 
```

## Exercise 3: Mapping Variables to Aesthetics

Create a scatterplot of `Sepal.Length` vs `Petal.Width` where:

 - Points are colored by `Species` 

 - Points are sized by `Sepal.Width` 

 - Add a title: "Iris Measurements by Species"
 
 - Use a clean theme (e.g., `theme_minimal()`)

```
# Your code here:
```

## Solution

```{r, fig.width=9, fig.height=5}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Width, color = Species, size = Sepal.Width)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Iris Measurements by Species",
    x = "Sepal Length (cm)",
    y = "Petal Width (cm)",
    color = "Species",
    size = "Sepal Width (cm)"
  ) +
  theme_minimal()
```

## Exercise 4: Bar Chart with Customization

Create a horizontal bar chart showing the count of flowers by `Species`:

 - Bars should be filled with "darkgreen"

 - Bars should have black borders (`color = "black"`)

 - Add count labels on top of each bar

 - Title: "Number of Flowers by Iris Species"

 - Use `coord_flip()` to make it horizontal

```
# Your code here:
```

## Solution

```{r, fig.width=9, fig.height=6}
iris %>%
  count(Species) %>%
  ggplot(aes(x = reorder(Species, n), y = n)) +
  geom_bar(stat = "identity", fill = "darkgreen", color = "black") +
  geom_text(aes(label = n), hjust = -0.2, size = 5) +
  coord_flip() +
  labs(
    title = "Number of Flowers by Iris Species",
    x = "",
    y = "Count"
  ) +
  theme_minimal()
```

## Exercise 5: Histogram with Density Curve

Create a histogram of `Sepal.Width` with:

 - 15 bins

 - Light blue fill with white borders

 - Overlay a density curve (`geom_density()`)

 - Facet by `Species` in 3 columns

 - Title: `"Sepal Width Distribution by Iris Species"`

```
# Your code here:
```

## Solution

```{r, fig.width=9, fig.height=5}
ggplot(iris, aes(x = Sepal.Width, y = ..density..)) +
  geom_histogram(fill = "lightgreen", color = "white", bins = 15, alpha = 0.7) +
  geom_density(color = "darkgreen", size = 1.2) +
  facet_wrap(~Species, ncol = 3) +
  labs(
    title = "Sepal Width Distribution by Iris Species",
    x = "Sepal Width (cm)",
    y = "Density"
  ) +
  theme_minimal()
```

## Exercise 6: Boxplot by Category

Create a boxplot showing the distribution of `Petal.Length` for each Species:

- Fill the boxes by Species with the colors: `"#66C2A5" (setosa)`, `"#FC8D62" (versicolor)`, `"#8DA0CB" (virginica)`

- Add a title: "Petal Length Distribution by Iris Species"

- Add an informative subtitle

```
# Your code here:
```

## Solution

```{r, fig.width=9, fig.height=5}
ggplot(iris, aes(x = Species, y = Petal.Length, fill = Species)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red", outlier.size = 3) +
  scale_fill_manual(values = c("setosa" = "#66C2A5", 
                                "versicolor" = "#FC8D62", 
                                "virginica" = "#8DA0CB")) +
  labs(
    title = "Petal Length Distribution by Iris Species",
    subtitle = "Virginica has the longest petals, setosa the shortest",
    x = "Species",
    y = "Petal Length (cm)",
    fill = "Species"
  ) +
  theme_minimal()
```

## Exercise 7: Faceting with `facet_wrap()`

Create a scatterplot of `Sepal.Length` vs `Petal.Length`:
  
  - Color points by `Species`

- Add a linear regression line (`method = "lm"`) for each panel

- Use `facet_wrap()` to create panels by Species in 1 row

- Title: "Relationship Between Sepal Length and Petal Length by Species"

```
# Your code here:
```

## Solution

```{r, fig.width=9, fig.height=7}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point(alpha = 0.7, size = 3) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  facet_wrap(~Species, nrow = 1) +
  labs(
    title = "Relationship Between Sepal Length and Petal Length by Species",
    x = "Sepal Length (cm)",
    y = "Petal Length (cm)",
    color = "Species"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Exercise 8: Stacked vs. Grouped Bar Chart Comparison

**Part A:** Create a **stacked bar chart** (this will be less meaningful with iris since Species is the only categorical variable; instead, create a new categorical variable `Size_Category` based on Sepal.Length).

First, create a size category variable:

```{r, fig.width=11, fig.height=5}
iris_with_size <- iris %>%
  mutate(Size_Category = case_when(
    Sepal.Length < 5 ~ "Small",
    Sepal.Length < 6 ~ "Medium",
    TRUE ~ "Large"
  ))
```

Then create a **stacked bar chart** showing count by `Size_Category`, with fill representing `Species`.

**Part B:** Create a **grouped (dodged) bar chart** with the same variables. 

```
# Part A: Your code here:

# Part B: Your code here:
```

## Solution

```{r, fig.width=11, fig.height=5}
# Create the data with size categories
iris_with_size <- iris %>%
  mutate(Size_Category = case_when(
    Sepal.Length < 5 ~ "Small",
    Sepal.Length < 6 ~ "Medium",
    TRUE ~ "Large"
  ))

# Part A: Stacked bar chart
p_stacked <- ggplot(iris_with_size, aes(x = Size_Category, fill = Species)) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("setosa" = "#66C2A5", 
                                "versicolor" = "#FC8D62", 
                                "virginica" = "#8DA0CB")) +
  labs(
    title = "Stacked Bar Chart: Species Composition by Sepal Length Category",
    x = "Sepal Length Category",
    y = "Count",
    fill = "Species"
  ) +
  theme_minimal()

# Part B: Grouped bar chart
p_dodged <- ggplot(iris_with_size, aes(x = Size_Category, fill = Species)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("setosa" = "#66C2A5", 
                                "versicolor" = "#FC8D62", 
                                "virginica" = "#8DA0CB")) +
  labs(
    title = "Grouped Bar Chart: Species Comparison by Sepal Length Category",
    x = "Sepal Length Category",
    y = "Count",
    fill = "Species"
  ) +
  theme_minimal()

# Display both (requires patchwork package)
library(patchwork)
p_stacked + p_dodged
```

## Exercise 9: Complete Data Visualization Pipeline

Using the `iris` dataset, create a comprehensive visualization that:
  
- Filters to only `setosa` and `versicolor` species

- Creates a scatterplot of `Sepal.Length` vs `Sepal.Width`

- Colors points by `Species`

- Uses different point shapes based on `Species`

- Adds a smooth trend line (`LOESS` not linear) without standard error

- Facets by `Species` in 2 columns

- Adds a meaningful title, subtitle, and axis labels

```
# Your code here:
```

## Solution

```{r, fig.width=9, fig.height=5}
iris %>%
  filter(Species %in% c("setosa", "versicolor")) %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color = Species, shape = Species)) +
  geom_point(size = 3.5, alpha = 0.7) +
  geom_smooth(se = FALSE, method = "loess", size = 1) +
  facet_wrap(~Species, ncol = 2) +
  labs(
    title = "Sepal Dimensions by Iris Species",
    subtitle = "Setosa shows a different sepal shape pattern compared to Versicolor",
    x = "Sepal Length (cm)",
    y = "Sepal Width (cm)",
    color = "Species",
    shape = "Species"
  ) +
  scale_color_manual(values = c("setosa" = "#66C2A5", "versicolor" = "#FC8D62")) +
  scale_shape_manual(values = c(16, 17)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Exercise 10: Advanced Customization Challenge

Create a publication-ready plot that answers the question: **"How does the relationship between sepal length and petal length differ by iris species?"**

Requirements:
  
- Scatterplot with `Sepal.Length` on x-axis, `Petal.Length` on y-axis

- Color by `Species`

- Add linear regression lines for each species with confidence intervals

- Add a separate panel for each species using faceting

- Customize colors using a meaningful palette

- Add a title and subtitle

- Add caption: "Data Source: Edgar Anderson's Iris Dataset (1935)"

- Apply a professional theme

- Add axis labels with units

```
# Your code here:
```

## Solution

```{r, fig.width=9, fig.height=8}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, size = 1) +
  facet_wrap(~Species, ncol = 3, scales = "free") +
  labs(
    title = "Relationship Between Sepal Length and Petal Length",
    subtitle = "Each iris species shows a distinct linear pattern with different slopes",
    x = "Sepal Length (cm)",
    y = "Petal Length (cm)",
    caption = "Data Source: Edgar Anderson's Iris Dataset (1935)",
    color = "Species"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.background = element_rect(fill = "lightgray"),
    strip.text = element_text(face = "bold", size = 11),
    plot.caption = element_text(hjust = 0, face = "italic")
  )
```