---
title: "__Topic 4: R Markdown__"
author: "Esam Mahdi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

# Introduction 

**R Markdown** (`.Rmd`) is a file format that combines three things in one document:

- **Text** (your explanations, answers, interpretations)

- **R** Code (your actual **R** commands)

- **Results** (the output from your code, including numbers, tables, and plots)

When you create an **R Markdown** file, you write everything in plain text. Then you **render** (or **"knit"**) the file, and **R** produces a beautiful final document showing your text, your code, and your results all together.

**Why R Markdown is recommended to use?**

| Reason | Explanation |
|--------|-------------|
| **Readability** | Anyone can read your explanations in plain English, see your code, and view your results all in one place—no more guessing. |
| **Reproducibility** | Anyone can re-run your entire file to verify that your code works and produces the same results. |
| **Organization** | Each task or question is clearly structured with explanation, code, and results in one place. |
| **Professional standard** | **R Markdown** is widely used by **data scientists** for reports, helping you follow industry best practices. |


## The Three Components of an R Markdown File

Every **R Markdown** (`.Rmd`) file has three essential parts. Here they are in order:

### 1. The YAML Header (Metadata)

The **YAML** header is at the **very top** of the file. It is surrounded by three dashes `---` on the top and bottom.

Here is a minimal, default YAML header:
yaml

```
---
title: "Title of Your Document"
author: "Your Name"
date: "2026-04-26"
output: html_document
---
```

**What each part does:**

| Element | What it does | What you must do |
|---------|--------------|------------------|
| `title` | Sets the title of your document | **Always change this** to the document name (e.g., "Homework 1") |
| `author` | Shows who wrote the document | **Always change this** to your full name |
| `date` | Shows the date | Keep as `"r Sys.Date()"` for auto-date, or write a specific date |
| `output` | Determines the final file format | Use `html_document` (see note below) |

**Important Note: You Can Knit to PDF or Word Too!**

Even though we use `output: html_document` in the header, you can still knit your `.Rmd` file to **PDF** or **Word** formats. Here is how:

| Output Format | How to Get It | When to Use |
|---------------|---------------|-------------|
| **HTML** | Click the "Knit" button (default) | Best for general submission; works everywhere |
| **PDF** | In the YAML header, change `output: html_document` to `output: pdf_document` | When requires a PDF |
| **Word** | In the YAML header, change `output: html_document` to `output: word_document` | When you need to edit in Microsoft Word |

**To knit to PDF or Word:**

1. Change the `output:` line in your YAML header

2. Click the `"Knit"` button

3. **R** will create the requested file type

**Example YAML headers for other formats:**

```
# For PDF output
output: pdf_document

# For Word output
output: word_document
```

**Note:** To knit to PDF, you may need to install LaTeX (RStudio will prompt you if it’s missing). You can also install it manually using TinyTeX: [https://yihui.org/tinytex/](https://yihui.org/tinytex/). For Word output, no additional installation is required.

### 2. Markdown Text (Your Explanations)

Everything in the file that is not inside a code chunk is written in **Markdown**. **Markdown** is a simple way to format plain text without using a mouse or menus.

**Basic Markdown You Will Use Every Time:**

| What you want | Type this | Result |
|---------------|-----------|--------|
| Big heading | `# Question 1` | **Question 1** (large and bold) |
| Smaller heading | `## Part (a)` | **Part (a)** (medium) |
| Even smaller heading | `### Step 1` | **Step 1** (smaller) |
| Italic text | `*explanation*` or `_explanation_` | *explanation* |
| Bold text | `**important**` or `__important__` | **important** |
| Bullet points | `- first item`<br>`- second item` | • first item<br>• second item |
| Numbered list | `1. step one`<br>`2. step two` | 1. step one<br>2. step two |
| Line break | Add two spaces at end of line then press Enter | New line starts |
| Code inside a sentence | \``mean(x)`\` | `mean(x)` |

**Writing Mathematical Formulas with LaTeX**

You can write mathematical formulas using **LaTeX** (a typesetting system for math). This is very useful for statistics and data science assignments.

**How to use LaTeX in R Markdown:**

| What you want | Type this | Rendered result |
|------------------------------|----------------------------|-----------|
| Inline formula (inside a sentence) | `$\bar{x} = 5$` | $\bar{x} = 5$ |
| Display formula (centered on its own line) | `$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$` | $$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$ |

**Common LaTeX symbols for this course:**

| Symbol | LaTeX code | Example |
|--------|------------|---------|
| Mean (x-bar) | `\bar{x}` | $\bar{x}$ |
| Summation | `\sum` | $\sum$ |
| Subscript | `x_i` | $x_i$ |
| Fraction | `\frac{a}{b}` | $\frac{a}{b}$ |
| Greek mu | `\mu` | $\mu$ |
| Greek sigma | `\sigma` | $\sigma$ |
| Square root | `\sqrt{x}` | $\sqrt{x}$ |
| Plus/minus | `\pm` | $\pm$ |

**Example of writing a formula in your answer:**

```
The formula for the sample mean is:

$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$

Where:
- $\bar{x}$ is the sample mean
- $x_i$ is each individual value
- $n$ is the number of values
```

This produces a professionally formatted mathematical explanation.

### 3. R Code Chunks (Your R Commands)

A **code chunk** is a box where you write R code. When you knit the document, R runs that code and shows the output right below it.

**Basic Code Chunk Syntax:**

Start a code chunk with three backticks and `{r}`. Then close the code chunk with three backticks. Then write your R code inside

````
```{r 
Write your R code inside
``` 
````

**Example of a Code Chunk:**

````
```{r 
my_numbers <- c(5, 10, 15, 20)
average <- mean(my_numbers)
average
```
````

When you knit the document, this will show:

  - The **R** code (because `echo = TRUE` is default)

  - The output: `[1] 12.5`

# Code Chunk Options (Very Important)

You can add options inside the curly braces `{r}` to control what appears in the final document.

| Option | What it does | When to use it |
|--------|--------------|----------------|
| `echo=TRUE` | Shows the R code (this is the default) | When you want to show your code |
| `echo=FALSE` | Hides the R code, shows only the output | For the **final answer** when you just need to show the result |
| `include=FALSE` | Hides both code and output | For setup code that loads packages or prepares data |
| `eval=FALSE` | Shows the code but does not run it | To show code as an example |

## Examples of Chunk Options:

### Example 1: Show code and output (default)

````
```{r 
sum(1:10)
```
````

### Example 2: Hide code, show only output (for final answer)

````
```{r echo=FALSE
sum(1:10)
```
````

This shows only: [1] 55

### Example 3: Hide everything (for setup)

````
```{r include=FALSE}
library(dplyr)
library(ggplot2)
```
````
### The Setup Chunk (Always Include This)

At the top of your **R Markdown** file, right after the YAML header, always include this setup chunk:

````
```{r setup, include=FALSE
knitr::opts_chunk$set(echo = TRUE)
```
````

This tells **R** to **show code by default** for all chunks. You can still use `echo=FALSE` for specific chunks when you want to hide code.

# Customizing Your Report Appearance

## R Markdown Themes

**R Markdown** HTML documents support **themes** that change the visual appearance (colors, fonts, spacing) of your report. You do not need to install anything extra – themes are built into **R Markdown**.

**Popular Themes from Bootswatch**

The themes come from [Bootswatch](https://bootswatch.com/), a collection of free themes for Bootstrap (the framework behind **R Markdown HTML**). Here are the most popular ones:

| Theme | Description | Best for |
|-------|-------------|----------|
| `cerulean` | Clean, blue accents, default-like | General purpose (very popular) |
| `cosmo` | Modern, sleek, slightly rounded corners | Contemporary reports |
| `flatly` | Minimal, flat design, professional | Business/formal reports |
| `journal` | Newspaper-style, serif font | Academic notes, papers |
| `readable` | High contrast, larger text | Teaching materials, accessibility |
| `darkly` | Dark background, light text | Working in low-light environments |

**How to Apply a Theme**

In your YAML header, add a `theme`: line under `html_document`:

```
---
title: "My Themed Report"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
---
```

To try a different theme, replace `cerulean` with any theme name: `cosmo, flatly, journal, readable, or darkly`.

**Example with multiple options:**


```
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
```

**Note:**  
- `toc: true` adds a table of contents to your document.  
- `toc_float: true` makes the table of contents stay visible while you scroll.

**Do I Have to Use a Theme?**

No. If you do not specify a theme, **R Markdown** uses a default style. However, using a theme makes your reports look more professional and easier to read.

## Your Complete Workflow for Submitting Assignments

**Step 1: Create a New R Markdown File**

  - Open RStudio

  - Click `File → New File → R Markdown...`

  - A dialog box appears. For now, just click `OK` (accept the defaults)

**RStudio** creates a new file with example content. **Delete all of that example content** except the YAML header.

**Step 2: Write Your YAML Header**

Make sure your header looks like this (change the title, author, and theme):

```
---
title: "Homework 1: Basic Calculations"
author: "Esam Mahdi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
---
```

**Step 3: Add the Setup Chunk**

Immediately after the YAML header, add the setup chunk:

````
```{r setup, include=FALSE
knitr::opts_chunk$set(echo = TRUE)
```
````

**Step 4: Write Your First Question**

Use markdown headings and text, then a code chunk:

````
# Question 1

Calculate the sum of the numbers from 1 to 10.

```{r
sum(1:10)
```
````

**Step 5: Write Your Second Question (with LaTeX)**

You can include mathematical formulas:

````
# Question 2

The formula for the sample mean is:

$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$

Calculate the mean of the numbers 2, 4, 6, 8, 10.

```{r
numbers <- c(2, 4, 6, 8, 10)
mean(numbers)
```
````

**Step 6: For Final Answers, Hide the Code**

If a question asks for just the final answer, use `echo=FALSE`:

````
# Question 3

What is the result of 15 multiplied by 3?

```{r echo=FALSE
15 * 3
```
````

This shows only `[1] 45` (the answer) without showing the multiplication code.

**Step 7: Knit to HTML (The Most Important Step)**

**Before you submit anything, you must click the KNIT button.**

- Look for the blue yarn icon with the word **"Knit"** above your script editor.

- Click it.

- **RStudio** will create an HTML file in the same folder as your `.Rmd` file.

- A preview will open in your browser.

**Check the HTML file carefully:**

- Is your name on it?

- Is the theme applied correctly?

- Do your LaTeX formulas look correct?

- Does it show all your answers?

- Are there any errors?

If something looks wrong, fix your `.Rmd` file and knit again.

**Step 8: Submit Your Files**

Your instructor will tell you which files to submit. Usually you submit:

- The `.Rmd` file (your source code)

- The `.html` file (the knitted report)

**Never submit a file that you have not knitted successfully.**

# Complete Example: A Full Assignment Solution

Here is a complete, simple **R Markdown** file answering three questions with a theme and LaTeX:


````
---
title: "Homework 1: Basic R Calculations"
author: "Esam Mahdi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
---

```{r setup, include=FALSE
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

Add the numbers 25 and 17.

```{r
25 + 17
```

# Question 2

The formula for the sample mean is:

$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$

Create a vector containing 3, 6, 9, 12, 15. Then calculate the mean of these numbers.

```{r
my_vector <- c(3, 6, 9, 12, 15)
mean(my_vector)
```

# Question 3

Calculate the area of a circle with radius 5. Use the formula: $A = \pi r^2$.  
(Hint: Use `pi` in R)

**Final answer only (code hidden):**

```{r echo=FALSE
pi * 5^2
```
````

When you knit this file, you get a clean HTML document with:

- The **cerulean** theme applied

- Your name and date at the top

- Three clearly labeled questions

- A properly formatted LaTeX formula for the mean

- Code shown for questions 1 and 2

- Only the answer shown for question 3

# Common Mistakes and How to Avoid Them

| Mistake | Why It Is a Problem | Solution |
|--------------------------|--------------------|----------------------------|
| Forgetting to change `author` in YAML | Your assignment has no name | Always check header before knitting |
| No empty line between chunks and text | Knitting may fail | Always put a blank line between markdown and code chunks |
| Missing the closing ` ``` ` | Knitting fails completely | Count your backticks - each chunk needs three at start and three at end |
| Syntax error in R code (e.g., missing parenthesis) | Knitting fails or output is wrong | Run each chunk with `Ctrl+Enter` before knitting |
| Forgetting to knit before submitting | You submit a broken file | **Always knit and check the HTML** |
| No explanations, just code chunks | Instructor sees no reasoning | Write at least one sentence in markdown before each code chunk |
| LaTeX syntax error (e.g., missing `$`) | Formula does not render correctly | Check that inline math has `$` on both sides |
| Wrong theme name in YAML | Knitting may fail or use default theme | Use exact theme names: `cerulean`, `cosmo`, `flatly`, etc. |

# Summary

| Concept | Key Point |
|---------|-----------|
| **What is R Markdown?** | A file format that combines text, code, and results |
| **File extension** | `.Rmd` |
| **Three components** | YAML header + Markdown text + R code chunks |
| **YAML header** | At top between `---` lines. Contains title, author, date, output format |
| **Themes** | Add `theme:` under `html_document` (e.g., `cerulean`, `cosmo`, `flatly`) |
| **Output formats** | `html_document` (default), `pdf_document`, `word_document` |
| **LaTeX math** | Use `$formula$` for inline, `$$formula$$` for display |
| **Markdown** | Simple formatting: `#` for headings, `*` for italics, `**` for bold |
| **Code chunk** | Starts with ` ```{r} ` and ends with ` ``` ` |
| **Default chunk option** | `echo=TRUE` (shows both code and output) |
| **Hide code** | Use `echo=FALSE` inside the curly braces |
| **Knit** | Click the Knit button to create the final report |
| **Always do this** | Knit before submitting and check the output |

# Exercises

## Exercise 1: Identify the Parts

Look at the **R Markdown** code below. Identify:

1. The YAML header

2. The theme being used

3. A markdown heading

4. A code chunk

5. A LaTeX formula

````
---
title: "Practice with Theme"
author: "Student"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
---

# Question 1

The mean formula is $\bar{x} = \frac{\sum x_i}{n}$.

```{r
mean(c(1, 2, 3, 4, 5))
```
````
## Solution

| Question | Answer |
|----------|--------|
| 1. YAML header | Lines 1-9 (from first `---` to `---`) |
| 2. Theme being used | `cosmo` |
| 3. Markdown heading | `# Question 1` |
| 4. Code chunk | Lines 13-15 (from `` ````{r}`` to ` ``` `) |
| 5. LaTeX formula | `$\bar{x} = \frac{\sum x_i}{n}$` |

## Exercise 2: Create an R Markdown File with Theme Cerulean and LaTeX

**Task**: Create a complete **R Markdown** file that:

1. Uses the cerulean theme in the YAML header

2. Includes the LaTeX formula $\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$ (mean)

3. Creates a vector data <- c(4, 8, 12, 16, 20)

4. Calculates and displays the mean of the vector

5. Shows the code and the output

**Requirements:**

- Change the title to "Exercise 2: Mean Calculation with Theme"

- Change the author to your name

- Write a brief markdown explanation before the code

- Use a markdown heading

**What to submit:** Copy and paste your entire `.Rmd` file content as your answer.

## Solution

Your `.Rmd` file should look like this:

````
---
title: "Exercise 2: Mean Calculation with Theme"
author: "Your Name Here"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
---

```{r setup, include=FALSE
knitr::opts_chunk$set(echo = TRUE)
```

# Calculating the Mean

The sample mean is calculated using the formula:

$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$

Below, I create a vector of numbers and compute its mean.

```{r
data <- c(4, 8, 12, 16, 20)
mean(data)
```
````

## Exercise 3: Fix the Errors

This **R Markdown** file has **errors**. Find and fix them.

````
---
title: "Broken Assignment"
author: "Student"
output: html_document:
  theme cerulean
---

```{r setup, include=FALSE
knitr::opts_chunk$set(echo = TRUE)

# Question 1

Calculate 10 divided by 2

```{r
10 / 2
```

# Question 2

The formula for the mean is $\bar{x} = \frac{\sum x_i}{n}

```{r echo=FALSE
mean(c(1, 2, 3)
```
````

List each error and write the correction.

## Solution

**Errors identified:**

| Error | Correction |
|-------|------------|
| YAML syntax: `output: html_document:` on same line as `theme` | Move `output: html_document` to its own line, indent `theme: cerulean` below it |
| YAML syntax: `theme cerulean` missing colon | Change to `theme: cerulean` |
| Setup chunk missing closing ` ``` ` | Add ` ``` ` after the chunk |
| No blank line between setup chunk and `# Question 1` | Add a blank line |
| LaTeX formula missing closing `$` | Change `$\bar{x} = \frac{\sum x_i}{n}` to `$\bar{x} = \frac{\sum x_i}{n}$` |
| Code chunk missing closing `}` in header | Change `{r echo=FALSE` to `{r echo=FALSE}` |
| Code chunk missing closing parenthesis | Change `mean(c(1, 2, 3)` to `mean(c(1, 2, 3))` |

**Corrected version:**

````
---
title: "Broken Assignment"
author: "Student"
output:
  html_document:
    theme: cerulean
---

```{r setup, include=FALSE
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

Calculate 10 divided by 2

```{r
10 / 2
```

# Question 2

The formula for the mean is $\bar{x} = \frac{\sum x_i}{n}$

```{r echo=FALSE
mean(c(1, 2, 3))
```
````

## Exercise 4: Try Different Output Formats

Write the YAML headers for three different versions of the same document:

1. HTML output with the `flatly` theme

2. PDF output (no theme needed)

3. Word output (no theme needed)

## Solution

| Output Format | YAML Header |
|---------|-------------------|
| HTML with flatly theme | `output:`<br>`  html_document:`<br>`    theme: flatly` |
| PDF output | `output: pdf_document` |
| Word output | `output: word_document` |

## Exercise 5: LaTeX Formula Practice

Write the **R Markdown** code (just the markdown/LaTeX part, not the YAML or R code) to display:

1. The formula for variance: $s^2 = \frac{\sum_{i=1}^{n} (x_i - \bar{x})^2}{n-1}$ (displayed on its own line)

2. The Greek letter mu ($\mu$) inline in a sentence

3. A fraction: $\frac{a+b}{c+d}$ (inline)

## Solution

| Task | R Markdown Code |
|------|-----------------|
| Variance formula (displayed) | `$$s^2 = \frac{\sum_{i=1}^{n} (x_i - \bar{x})^2}{n-1}$$` |
| Greek mu inline | `The population mean is denoted by $\mu$.` |
| Fraction inline | `The ratio $\frac{a+b}{c+d}$ is greater than 1.` |



