---
title: "__Topic 1: From R Installation to Basic Calculations__"
author: "Esam Mahdi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    self_contained: true
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

# Getting Started: Installing R and RStudio

Before you can run any **R** code, you need to install two things:

1. **R** (the engine): Download it from [CRAN (Comprehensive R Archive Network](https://cran.r-project.org/). Choose your operating system (Windows, Mac, or Linux) and follow the default installation steps.

2. **RStudio** (the interface): Download the free Desktop version from [posit.co](https://posit.co/download/rstudio-desktop). Install it after installing **R**. **RStudio** will automatically detect your **R** installation.

Once both are installed, launch **RStudio** (not **R** directly).

## Why Switch from R to RStudio?

You could write **R** code directly in the basic **R** console that opens when you click the **R** icon. However, **RStudio** provides a much more productive environment.

**Forget using plain R. Use RStudio from now on because:**

- It offers **four organized panes** (Script Editor, Console, Environment/History, Files/Plots/Packages/Help).

- You can write and save **scripts** (`.R` or `.Rmd` files) to reuse your work.

- It has **syntax highlighting**, code completion, and built-in plotting viewer.

- It integrates **R Markdown**, version control (`Git`), and project management.

All the code in this document can be run in RStudio by clicking the green **"Run"** button or pressing `Ctrl+Enter` (Windows/Linux) / `Cmd+Enter` (Mac) on a line or selected block.

## Understanding R Packages: Base, Main, and Outsider

R's functionality is organized into **packages** (collections of functions, data, and documentation). When you install R, you automatically receive two distinct sets of packages: **Base packages** and **Recommended (Main) packages**. Anything beyond these is called an **Outsider (Contributed) package**.

### Base Packages (Always Available, No `library()` Needed)

Base (essential) packages are the absolute core of **R**. They are loaded automatically when you start **R**. You **do not** need to use `library()` for them. There are 15 base packages. Here is the complete list of Base packages in a standard **R** installation:

| Package | Description |
|-------|-------------------------|
| `base` | Fundamental **R** functions: arithmetic, `I/O`, data structures, condition handling, etc. |
| `compiler` | Byte code compiler for **R** functions. |
| `datasets` | Built-in example datasets (e.g., `iris`, `mtcars`, `AirPassengers`). |
| `graphics` | Base graphics functions for plotting (e.g., `plot()`, `hist()`, `barplot()`). |
| `grDevices` | Graphics device management and color handling (e.g., `colors()`, `png()`, `pdf()`). |
| `grid` | A more flexible graphics system than base graphics (alternative to graphics). |
| `methods` | Formal object-oriented programming (S4 classes and methods). |
| `parallel` | Support for parallel computing (`snow`, `multicore`, etc.). |
| `splines` | Regression spline functions and classes. |
| `stats` | Extensive statistical functions: distributions, tests, linear models, clustering, etc. |
| `stats4` | S4 classes and methods for statistical modeling. |
| `tcltk` | Interface to the Tcl/Tk GUI toolkit (for creating simple GUI windows). |
| `translations` | A system package containing pre-compiled translation files (`.mo` files) used to display R’s messages and errors in different languages. |
| `tools` | Tools for package development and checking. |
| `utils` | Utilities: package installation, data entry, file handling, object inspection (`str()`, `head()`, `summary()`). |
  
**Note:** `methods` and `stats4` are technically part of the base distribution but may require explicit loading with `library(methods)` in some very old or minimal **R** setups. However, in modern **R**, they load automatically.

### Main (Recommended) Packages

These are Recommended packages distributed with **R** but **not loaded by default**. You must load them using `library(package_name)` when you need them. They are maintained by the **R** Core Team or highly trusted developers. There are 15 recommended packages. 

## Main (Recommended) Packages

| Package | Description |
|-------|---------------|
| `boot` | Bootstrap functions (resampling methods) and related data. |
| `class` | Functions for classification (k-nearest neighbours, LVQ, etc.). |
| `cluster` | Clustering algorithms (hierarchical, k-means, PAM, etc.). |
| `codetools` | Code analysis tools for R (finding unused variables, detecting issues). |
| `foreign` | Read and write data from other statistical systems (SPSS, Stata, SAS, etc.). |
| `KernSmooth` | Kernel smoothing and density estimation (Wand & Jones methods). |
| `lattice` | Trellis graphics (multivariate data visualization, an alternative to ggplot2). |
| `MASS` | Functions from the book "Modern Applied Statistics with S" (Venables & Ripley). |
| `Matrix` | Sparse and dense matrix operations and classes. |
| `mgcv` | Generalized additive models (GAMs) with smooth term estimation. |
| `nlme` | Linear and nonlinear mixed-effects models. |
| `nnet` | Feed-forward neural networks and multinomial log-linear models. |
| `rpart` | Recursive partitioning and regression trees (CART-like models). |
| `spatial` | Functions for spatial statistics (Kriging, point patterns). |
| `survival` | Survival analysis (Kaplan-Meier, Cox proportional hazards models). |

### Outsider (Contributed) Packages

These are created by the global **R** community and hosted on CRAN, Bioconductor, or GitHub. They cover almost everything: data manipulation (`dplyr`), visualization (`ggplot2`), machine learning (`caret, tidymodels`), reporting (`rmarkdown`), web apps (`shiny`), and thousands more.

**You must install outsider packages first before using them** (only once per **R** installation):

```
# Install a single package from CRAN
install.packages("ggplot2")

# Install multiple packages at once
install.packages(c("dplyr", "tidyr", "readr", "purrr"))
```

After installation, load the package into your current **R** session using `library()` or `require()`:

```
library(ggplot2)
library(dplyr)
```

**How to See Which Packages Are Already Installed (Base + Main + Outsiders)**

```
# See all installed packages (Base + Main + any outsiders you added)
installed.packages()

# See only Base packages
installed.packages()[installed.packages()[, "Priority"] == "base", "Package"]

# See only Recommended (Main) packages
installed.packages()[installed.packages()[, "Priority"] == "recommended", "Package"]
```

**Important distinction:**

- **Base packages** → Always available. No installation. No `library()` or `require()`` needed.

- **Main (Recommended) packages** → Already installed with **R**. Need `library()` to use.

- **Outsider packages** → Need `install.packages()` once, then `library()` each session.
    
# Variables

A **variable** is a named container that stores a value or object. Think of it like a labeled box where you can put data. Once you create a variable, you can use its name to access whatever value it holds.

**Why use variables?**

    - Store values for later use

    - Give meaningful names to data

    - Avoid repeating the same value multiple times

    - Make your code easier to read and maintain
    
## Creating Variables with `<-`

In **R**, the primary way to create a variable is using the assignment operator `<-` (a less-than sign followed by a hyphen).

**Basic Variable Assignment**
```{r}
# Store a number
age <- 25

# Store text
message <- "Learning R is fun!"

# Store the result of a calculation
is_equal <- (sqrt(144) == 12)

# Display the variables
age
message
is_equal
```

**Key Points:**

    - The variable name goes on the left.

    - The value goes on the right

    - The <- operator points from right to left (value goes into the variable)
  
    - The variable name can contain letters, numbers, dots (.), and underscores (_).

    - The variable name must start with a letter or a dot (but a dot cannot be followed by a number).

    - The variable name is case-sensitive (myVar is different from myvar).

    - The variable name cannot use reserved words like `if`, `else`, `TRUE`, `FALSE`

```{r}
# Valid variable names
student_age <- 20
student.age <- 21
student_age_2 <- 22
.student_age <- 23

# Invalid variable names (these would cause errors)
# 2student <- 20      # Can't start with a number
# student-age <- 20   # Hyphens not allowed (R thinks it's subtraction)
# if <- 20            # Reserved word
```

**Use Spaces for Readability**

Always put spaces around the assignment operator to make your code clearer:

```{r}
# Poor style (hard to read)
x<-5
x

# Good style (clear and readable)
x <- 5
x
```

**Variables Can Be Reassigned**

A variable always holds its most recent value. When you assign a new value, the old one is replaced:

```{r}
temperature <- 72
temperature  # Shows 72

temperature <- 68  # Reassign new value
temperature  # Now shows 68
```

**Using Variables in Calculations**

Once variables contain numeric values, you can use them in mathematical operations:

```{r}
width <- 10
height <- 5

area <- width * height
area  # 50

perimeter <- 2 * (width + height)
perimeter  # 30
```

**Note:** Many programming languages use `=` for assignment, but **R** prefers `<-`. Here's why:

**The Equals Sign Has Multiple Roles**

In **R**, `=` is used for two different purposes:

    Assigning function arguments (recommended)

    Assigning variables (possible but not recommended)

**Example:** When you use `=` inside a function call, it does NOT create a variable

```{r}
# Example: round() function has two arguments: x and digits
rounded <- round(x = 3.14159, digits = 2)  # 'digits = 2' sets a function argument and doesn't create a 'digits' variable
rounded
```

```
# Trying to access 'digits' causes an error
digits  # Would output: Error: object 'digits' not found
```

**The `<-` Operator Always Creates Variables**

```{r}
# Using <- inside a function creates a variable
round(3.14159, digits <- 2)  # Creates 'digits' variable
digits  # Now digits exists and has value 2
```

**Warning:** Creating variables inside function calls like this is usually NOT what you want. It can lead to unexpected behavior and cluttered workspaces.

**Best Practice Summary**

| Purpose	| Recommended Operator | Example
|---------------------|----------|-----------------|
| Variable assignment	| <- | my_var <- 10 |
| Function arguments	| =	|round(3.14, digits = 2) |
| Equality comparison |	== |	5 == 5 (returns TRUE) |

```{r}
# Correct usage examples
score <- 95                    # Variable assignment
result <- round(score / 3, digits = 2)  # Function argument with =

# Checking equality
score == 95                    # TRUE
score == 100                   # FALSE
```

## Alternative (to `<-`) Assignment Methods 

**The `=` Operator (Not Recommended)**

While you can use `=` for assignment, it's considered poor style:

```{r}
# Possible but not recommended
total = 100
total

# Better to use <-
total <- 100
```

**The `assign()` Function**

**R** also provides the `assign()` function for creating variables. This is useful in advanced programming scenarios:

```{r}
# assign(variable_name, value)
assign("apples", 5)
assign("oranges", 3)

apples
oranges

# Calculate total fruit
total_fruit <- apples + oranges
total_fruit  # 8
```

**Right Assignment `->` (Rarely Used)** 

**R** also allows assignment from left to right using `->`, but this is rarely used:

```{r}
# Right assignment (uncommon)
10 -> my_number
my_number  # 10

# Left assignment (standard and preferred)
my_number <- 10
```

# Examples

## Example 1: Tracking Personal Finances

```{r}
# Income and expenses
monthly_salary <- 5000
rent <- 1500
groceries <- 600
utilities <- 300
entertainment <- 400

# Calculate savings
total_expenses <- rent + groceries + utilities + entertainment
monthly_savings <- monthly_salary - total_expenses

monthly_savings  # 2200
```

## Example 2: Converting Temperatures

```{r}
# Convert Celsius to Fahrenheit
celsius_temp <- 25
fahrenheit_temp <- (celsius_temp * 9/5) + 32

fahrenheit_temp  # 77°F
```

## Example 3: Building a Simple Calculator

```{r}
# Store two numbers
num1 <- 15
num2 <- 4

# Perform operations
sum_result <- num1 + num2
difference <- num1 - num2
product <- num1 * num2
quotient <- num1 / num2

# Display results
sum_result      # 19
difference      # 11
product         # 60
quotient        # 3.75
```

# Common Mistakes to Avoid

**Mistake 1:** Using `<-` in a function argument when you meant `=`

```
# Wrong: round(3.14, digits <- 2)  # Creates unwanted variable
# Correct: round(3.14, digits = 2)  # This is correct as function argument
```

**Mistake 2:** Forgetting spaces

```
# Bad: x<-5
# Good: x <- 5
```

**Mistake 3:** Using `=` instead of `==` for comparison

```
# Wrong: if (x = 10)  # This is assignment, not comparison
# Correct: if (x == 10)  # This checks equality
```

**Mistake 4:** Using reserved words as variable names

```
# Don't: TRUE <- 5
# Don't: if <- 10
```

# Summary Table

| Concept | Operator | Example | Purpose |
|---------|----------|---------|---------|
| Variable assignment (preferred) | `<-` | `x <- 10` | Store values |
| Variable assignment (allowed but not preferred) | `=` | `x = 10` | Store values (avoid) |
| Function arguments | `=` | `round(x, digits = 2)` | Pass parameters to functions |
| Equality comparison | `==` | `x == 10` | Test if values are equal |
| Assignment from right to left (rare) | `->` | `10 -> x` | Alternative assignment |

# Exercises

## Exercise 1: Variable Assignment Order

What is the final value of result after running the code below? Write your answer, then uncomment and run the code to check.

```{r}
# Write your prediction here: ___________

# result <- 100
# result <- 50
# result <- 75
# result
```

## Solution

```{r}
# The final value is 75 (the last assignment)
result <- 100
result <- 50
result <- 75
result  # 75
```

## Exercise 2: Using Variables in Calculations

Create variables for length and width, then calculate the area of a rectangle. Replace the placeholders with your code.

```{r}
# Assign length as 12 and width as 8
# Then calculate area (length × width)

# Your code here:
# length <- 
# width <- 
# area <- 
# area
```

## Solution 
```{r}
length <- 12
width <- 8
area <- length * width
area  # 96
```

## Exercise 3: Temperature Conversion

Convert 100°F to Celsius using the formula: `°C = (°F - 32) × 5/9`

```{r}
# Store 100°F in a variable called fahrenheit
# Calculate celsius and store it in a variable
# Display the celsius value

# Your code here:
```

## Solution
```{r}
fahrenheit <- 100
celsius <- (fahrenheit - 32) * 5/9
celsius  # 37.77778
```

## Exercise 4: Understanding Assignment Operators

Predict what will happen when you run the code below. Then uncomment and run to check.

```{r}

# Prediction: Will 'digits' exist after this code runs?
# Why or why not?

# round(42.987, digits = 2)

# Check if 'digits' exists (uncomment to test)
# digits
```

## Solution 
```{r}
# digits will NOT exist because = inside function arguments
# does NOT create variables
round(42.987, digits = 2)
# digits  # Would give error: object 'digits' not found
```

## Exercise 5: Creating a Shopping Total

Create variables for different items and their prices, then calculate the total cost including 8% tax.

```{r}
# Item prices:
# Apples: $3.50
# Bananas: $2.00
# Oranges: $4.00
# Bread: $5.50

# Calculate subtotal, tax, and total
# Tax rate is 8% (0.08)
```

## Solution
```{r}
apples <- 3.50
bananas <- 2.00
oranges <- 4.00
bread <- 5.50

subtotal <- apples + bananas + oranges + bread
tax <- subtotal * 0.08
total <- subtotal + tax

total  # $16.20
```

## Exercise 6: Swapping Variables

Swap the values of two variables without directly reassigning them to their opposite values.

```{r}
a <- 10
b <- 20
# Write code to swap the values so a becomes 20 and b becomes 10
# Hint: Use a temporary variable

# Your solution here:
```

## Solution
```{r}
a <- 10
b <- 20

# Swap using temporary variable
temp <- a
a <- b
b <- temp

a  # 20
b  # 10
```

## Exercise 7

Install the `tidyverse` package, then load it and explain what you get.

*Hint: The `tidyverse` is actually a collection of multiple packages designed for data science. Use `install.packages()` to install it, then `library(tidyverse)` to load it. After loading, look at the message that appears in the console.*

```{r, eval=FALSE}
# Write your code here
```

1. How many packages are loaded when you run library(tidyverse)?

2. List at least 5 of the core packages included in the tidyverse.

3. What is the main purpose of the tidyverse?

## Solution

```
# Install and load tidyverse
install.packages("tidyverse")
library(tidyverse)

# Question 1 Answer:
# 8 packages (ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, forcats)

# Question 2 Answer:
# ggplot2 (visualization), dplyr (manipulation), tidyr (tidying), 
# readr (import), purrr (functional programming)

# Question 3 Answer:
# The tidyverse provides a consistent and efficient set of tools for 
# the entire data science workflow from importing to visualizing data.
```

