---
title: '__Topic 3: Vectors in R__'
author: "Esam Mahdi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to Vectors

In previous lectures, we worked with **single values** stored in variables. But real-world data rarely comes as isolated numbers. More often, you'll work with **collections** of related data: test scores for an entire class, daily temperatures for a month, or product prices in a shopping cart.

A **vector** is R's fundamental data structure for storing multiple values of the **same type** in a single container. Think of a vector as a row of connected boxes, each holding one value:

```
[ Box 1 ] [ Box 2 ] [ Box 3 ] [ Box 4 ] [ Box 5 ]
   25        30        35        40        45
```

**Why use vectors?**

    - Work with hundreds or thousands of values at once

    - Perform calculations on entire datasets with a single command

    - Filter and subset data based on conditions

    - Avoid writing tedious loops

**Key property:** All elements in a vector must be the same type (all numbers, all text, or all `TRUE/FALSE`).

## Creating Vectors
**The `c()` Function**

The most common way to create a vector is with `c()`, which stands for `"combine"` or `"concatenate"`. Place your values inside the parentheses, separated by commas.

```{r}
# Numeric vector (student ages)
ages <- c(18, 19, 22, 20, 21)
ages

# Character vector (product names)
products <- c("Laptop", "Mouse", "Keyboard", "Monitor")
products

# Logical vector (answers to a quiz)
quiz_answers <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
quiz_answers
```

**Creating Vectors with Patterns**

For large vectors, typing every value with `c()` is impractical. **R** provides several shortcuts:

The **colon operator** `:` $-$ creates sequences of whole numbers:

```{r}
# Numbers 1 through 15
numbers_1_to_15 <- 1:15
numbers_1_to_15

# Counting backwards
countdown <- 10:1
countdown
```

**The `seq()` function** $-$ gives you control over start, end, and step size:

```{r}
# Even numbers from 2 to 20
even_numbers <- seq(from = 2, to = 20, by = 2)
even_numbers

# Decimal sequence
decimals <- seq(from = 0, to = 1, by = 0.1)
decimals
```

**The `rep()` function** $-$ repeats values:

```{r}
# Repeat 0 five times
zeros <- rep(0, times = 5)
zeros

# Repeat a pattern
pattern <- rep(c("Yes", "No"), times = 4)
pattern
```

# Random Vectors

**R** can generate vectors of random numbers from various distributions:

```{r}
# 10 random numbers between 0 and 100 (uniform distribution)
random_scores <- runif(n = 10, min = 0, max = 100)
random_scores

# Round to whole numbers for easier reading
round(random_scores)
```

**The `set.seed()` Function**

Why do we need `set.seed()?`

Computers don't actually generate truly random numbers. Instead, they use mathematical algorithms to generate **pseudo-random numbers**. These algorithms start with a **seed value** (a starting number) and then produce a sequence that appears random.

**The problem:** Every time you run a random number generator, you could get a different sequence of numbers. This makes it difficult to:

  - Reproduce your results

  - Debug your code

  - Share reproducible research

**The solution:** `set.seed()` fixes the starting point of the random number generator so that you get the same `"random"` numbers every time you run your code.

**Example Without** `set.seed()`**:**

```{r}
# Run this code multiple times - you'll get different numbers each time!
runif(n = 5, min = 0, max = 10)
# Run again - different numbers!
runif(n = 5, min = 0, max = 10)
```

**Example With** `set.seed()`**:**

```{r}
# Set the seed to a specific number (any positive integer works)
set.seed(123)

# Generate random numbers
runif(n = 5, min = 0, max = 10)

# Reset the same seed
set.seed(123)

# Get the SAME "random" numbers again!
runif(n = 5, min = 0, max = 10)
```

**Note:** Different seed values produce different sequences of random numbers:

```{r}
set.seed(1)
runif(3, min = 0, max = 100)

set.seed(2)
runif(3, min = 0, max = 100)

set.seed(100)
runif(3, min = 0, max = 100)
```

## Examining Vectors

**Checking Length and Type**

Use these functions to inspect vectors:

```{r}
# Create a sample vector
prices <- c(9.99, 14.99, 24.99, 5.99)

# How many elements?
length(prices)

# What type of data?
typeof(prices)  # "double" means decimal numbers

# Quick summary
str(prices)  # Shows type and first few values
```

**Printing Vectors**

When you print a vector, **R** shows index numbers in brackets `[ ]` at the start of each line. The number tells you the position of the first value on that line:

```{r}
# Create 50 random numbers to see the index labels
big_vector <- runif(50)
big_vector  # Notice the [1], [7], [13], [19], [25], [31], [37], [43], [49] labels
```

These index labels help you understand where you are in a long vector.

## Accessing Vector Elements: Indexing

**The Golden Rule: R Starts at 1**

Unlike **Python**, **Java**, and **C++** (which start at 0), **R** starts indexing at 1. The first element is position 1, the second is position 2, and so on.

**Single Element Access**

Use square brackets `[]` after the vector name:

```{r}
# Create a vector of temperatures for a week (Monday = day 1)
week_temps <- c(72, 74, 68, 70, 75, 77, 73)
week_temps

# Get Wednesday's temperature (day 3)
week_temps[3]

# Get Sunday's temperature (day 7)
week_temps[7]
```

**Multiple Elements**

Range of consecutive elements with `::`

```{r}
# Get Monday through Wednesday (days 1-3)
week_temps[1:3]

# Get Friday through Sunday (days 5-7)
week_temps[5:7]
```

Specific non-consecutive elements with `c()`:

```{r}
# Get Monday, Wednesday, and Friday
week_temps[c(1, 3, 5)]
```

Excluding Elements with `-`

Use a minus sign to remove elements:

```{r}
# All days except Wednesday (remove index 3)
week_temps[-3]

# Remove first and last days
week_temps[-c(1, 7)]
```

**Invalid Indexes**

If you try to access a position that doesn't exist, **R** returns `NA` (Not Available):

```{r}
# Only 7 days in a week, so index 100 doesn't exist
week_temps[100]
```

## Logical Indexing: Filtering with Conditions

This is one of R's most powerful features. Instead of specifying which positions you want, you provide a condition (logical test), and **R** returns all elements that satisfy it.

### Step-by-Step Example

```{r}
# Test scores from 10 students
scores <- c(85, 92, 78, 65, 94, 88, 71, 96, 82, 79)

# Step 1: Create a logical vector (TRUE/FALSE) for condition
above_85 <- scores > 85
above_85  # TRUE at positions where score > 85

# Step 2: Use that logical vector to filter
high_scores <- scores[above_85]
high_scores
```

### One-Step Filtering

**R** allows you to combine everything in a single line:

```{r}
# Get all scores above 85 (direct)
scores[scores > 85]

# Get all scores below 70
scores[scores < 70]

# Get all scores between 80 and 90
scores[scores >= 80 & scores <= 90]
```

### Filtering Character Vectors

```{r}
# Student names and their grades
students <- c("Alice", "Bob", "Ahmed", "Diana", "Eve")
grades <- c("A", "C", "B", "A", "B")

# Get all students who got an A
students[grades == "A"]

# Get students who DIDN'T get an A
students[grades != "A"]
```

**The `%in%` Operator for Membership**

Use `%in%` to check if elements belong to a specified set:

```{r}
# Fruit inventory
fruits <- c("apple", "banana", "orange", "apple", "grape", "banana")

# Find all apples and oranges
fruits[fruits %in% c("apple", "orange")]
```

**Vectorized Operations: R's Superpower**

Vectorization means operations automatically apply to every element of a vector. You don't need loops like in other languages.

**Arithmetic with Single Values**

```{r}
# Price list
original_prices <- c(10, 20, 30, 40, 50)

# Apply 20% discount to ALL prices
discounted <- original_prices * 0.8
discounted

# Add tax to ALL prices (8%)
with_tax <- original_prices * 1.08
with_tax

# Round all prices to nearest dollar
rounded_prices <- round(with_tax)
rounded_prices
```

**Operations Between Two Vectors**

When you operate on two vectors of the same length, **R** pairs elements by position:

```{r}
# Hours worked by employees
hours <- c(40, 35, 38, 42, 36)

# Hourly wage for each employee
wage <- c(25, 30, 28, 35, 27)

# Calculate each employee's weekly pay
weekly_pay <- hours * wage
weekly_pay

# Calculate total payroll
total_payroll <- sum(weekly_pay)
total_payroll
```

**Vectorized Functions**

Many **R** functions work on entire vectors:

```{r}
# Square root of each number
numbers <- c(4, 9, 16, 25, 36)
sqrt(numbers)

# Round each number
decimals <- c(1.2, 2.7, 3.5, 4.1)
round(decimals)
floor(decimals)  # Round down
ceiling(decimals)  # Round up
```

**Character Vector Operations**

```{r}
# Combine first and last names element-wise
first_names <- c("John", "Jane", "Bob")
last_names <- c("Smith", "Doe", "Johnson")

full_names <- paste(first_names, last_names)
full_names

# Convert all names to uppercase
toupper(full_names)
```

**Combining Vectors**
Using `c()` to Join Vectors

```{r}
# Sales data for two weeks
week1_sales <- c(100, 150, 200)
week2_sales <- c(180, 220, 190)

# Combine for monthly report
month_sales <- c(week1_sales, week2_sales)
month_sales
```

**Type Coercion**

If you combine vectors of different types, **R** automatically converts everything to the most flexible type. The order of flexibility is: **character > numeric > logical**

```{r}
# Numeric + Character = Character
numbers <- c(1, 2, 3)
words <- c("A", "B", "C")
mixed <- c(numbers, words)
mixed  # Numbers become text
typeof(mixed)

# Logical + Numeric = Numeric (TRUE = 1, FALSE = 0)
logical_values <- c(TRUE, FALSE, TRUE)
numeric_values <- c(10, 20, 30)
combined <- c(logical_values, numeric_values)
combined
```

# Examples
## Example 1: Student Grade Analysis

```{r}
# Exam scores for a class of 15 students
exam_scores <- c(78, 85, 92, 67, 88, 74, 95, 81, 70, 86, 91, 77, 84, 89, 73)

# Basic statistics
mean_score <- mean(exam_scores)
min_score <- min(exam_scores)
max_score <- max(exam_scores)

cat("Class average:", mean_score, "\n")
cat("Highest score:", max_score, "\n")
cat("Lowest score:", min_score, "\n")

# Who passed? (score >= 70)
passing_scores <- exam_scores[exam_scores >= 70]
cat("Number of students who passed:", length(passing_scores), "\n")
cat("Passing scores:", passing_scores, "\n")

# Who got an A? (score >= 90)
a_scores <- exam_scores[exam_scores >= 90]
cat("A students' scores:", a_scores)
```

## Example 2: Weather Data Analysis

```{r}
# Daily high temperatures in July (31 days)
july_highs <- c(85, 87, 88, 90, 92, 91, 89, 88, 87, 86, 
                88, 90, 93, 95, 94, 92, 90, 89, 88, 87, 
                86, 85, 84, 86, 88, 90, 92, 93, 91, 89, 87)

# Days with extreme heat (over 90°F)
hot_days <- july_highs[july_highs > 90]
cat("Number of hot days (>90°F):", length(hot_days), "\n")

# Convert all temperatures to Celsius (formula: (°F - 32) × 5/9)
july_celsius <- (july_highs - 32) * 5/9
cat("Highs in Celsius (first 5 days):", round(july_celsius[1:5], 1), "\n")

# Weekly average (first week = days 1-7)
week1_avg <- mean(july_highs[1:7])
cat("Average temperature in week 1:", week1_avg, "°F\n")
```

## Example 3: E-commerce Shopping Cart

```{r}
# Shopping cart
cart_items <- c("Laptop", "Mouse", "Keyboard", "USB Cable", "Mouse Pad")
cart_prices <- c(899.99, 24.99, 49.99, 8.99, 12.99)

# Subtotal calculation
subtotal <- sum(cart_prices)
cat("Subtotal: $", subtotal, "\n")

# Apply 10% discount
discount <- subtotal * 0.10
after_discount <- subtotal - discount
cat("After 10% discount: $", after_discount, "\n")

# Add tax (8%)
tax <- after_discount * 0.08
final_total <- after_discount + tax
cat("Final total (with tax): $", round(final_total, 2), "\n")

# Find expensive items (over $50)
expensive_items <- cart_items[cart_prices > 50]
expensive_prices <- cart_prices[cart_prices > 50]
cat("Items over $50:", expensive_items, "\n")
cat("Their prices: $", expensive_prices)
```

# Common Mistakes
## Mistake 1: Mixing Types Accidentally

```{r}
# This works but might not be what you want
mixed_vector <- c(10, "20", 30)
mixed_vector  # All become text (notice the quotes)
typeof(mixed_vector)  # "character"
```

**Fix:** Keep vectors homogeneous. If you need different types, use lists (future lecture).

## Mistake 2: Off-by-One Errors

```{r}
my_vector <- c(10, 20, 30, 40, 50)

# This is wrong (indices start at 1, not 0)
my_vector[0]  # Returns empty vector, not 10

# Correct way to get first element
my_vector[1]  # Returns 10
```

## Mistake 3: Forgetting Parentheses with `c()`

```
# Wrong (creates a single number, not a vector)
wrong <- 1, 2, 3  # Syntax error

# Correct
correct <- c(1, 2, 3)
```

## Mistake 4: Using `=` for Equality Inside Brackets

```{r}
scores <- c(85, 92, 78, 65)

# Wrong (assignment, not comparison)
scores[scores = 92]  # Error (gives NA)

# Correct (comparison)
scores[scores == 92]  # Returns 92
```

# Summary

| Concept | Syntax | Example |
|---------|--------|---------|
| Create vector | `c()` | `c(-1, 20, 13)` |
| Sequence | `:` | `4:12` |
| Sequence with step | `seq()` | `seq(1, 10, 0.2)` |
| Repeat | `rep()` | `rep(2, times = 5)` |
| Get element | `vector[index]` | `scores[3]` |
| Get range | `vector[start:end]` | `scores[2:5]` |
| Exclude elements | `vector[-index]` | `scores[-1]` |
| Filter with condition | `vector[condition]` | `scores[scores > 80]` |
| Vector length | `length()` | `length(scores)` |
| Vector type | `typeof()` | `typeof(scores)` |
| Element-wise operation | `vector + scalar` | `prices * 0.8` |
| Element-wise between vectors | `vector1 + vector2` | `hours * wage` |

# Exercises
## Exercise 1: Create a Vector

Create a vector called temperatures containing the following Celsius temperatures: `22, 25, 21, 24, 26, 23, 20`. Then display the vector.

```{r}
# Your code here:
```

## Solution

```{r}
temperatures <- c(22, 25, 21, 24, 26, 23, 20)
temperatures
```

## Exercise 2: Vector Indexing

Using the temperatures vector from Exercise 1, extract:

    The 3rd temperature

    The last temperature

    All temperatures except the first two

```{r}
# Your code here:
```

## Solution 

```{r}
temperatures <- c(22, 25, 21, 24, 26, 23, 20)

# 3rd temperature
temperatures[3]

# Last temperature (can use length())
temperatures[length(temperatures)]

# All except first two
temperatures[-c(1, 2)]
# Alternative: temperatures[3:7]
```

## Exercise 3: Filtering with Conditions

You have the following test scores: `88, 72, 95, 68, 91, 77, 84, 63, 96, 82`. Write code to:

    Find all scores above 85

    Find all scores below 70

    Find all scores between 75 and 90 (inclusive)

```{r}
# Your code here:
```

## Solution 

```{r}
scores <- c(88, 72, 95, 68, 91, 77, 84, 63, 96, 82)

# Above 85
scores[scores > 85]

# Below 70
scores[scores < 70]

# Between 75 and 90 inclusive
scores[scores >= 75 & scores <= 90]
```

## Exercise 4: Vectorized Operations

A store is having a 15% off sale on all items. The original prices are: `25.99, 49.50, 12.75, 88.00, 34.25`. Calculate the sale price for each item and round to 2 decimal places.

```{r}
# Your code here:
```

## Solution 

```{r}
original_prices <- c(25.99, 49.50, 12.75, 88.00, 34.25)

# Apply 15% discount (multiply by 0.85)
sale_prices <- original_prices * 0.85

# Round to 2 decimal places
round(sale_prices, 2)
```

## Exercise 5: Combining and Summarizing

Daily steps for two weeks: `Week 1 = 8500, 7200, 9500, 6800, 10200, 8800, 7700. Week 2 = 9100, 8300, 7600, 10400, 8900, 7200, 9800`.

Combine both weeks into one vector and calculate:

    Total steps across all days

    Average daily steps

    Maximum steps in a single day

    Days with over 9000 steps

```{r}
# Your code here:
```

## Solution

```{r}
week1 <- c(8500, 7200, 9500, 6800, 10200, 8800, 7700)
week2 <- c(9100, 8300, 7600, 10400, 8900, 7200, 9800)

# Combine weeks
all_steps <- c(week1, week2)

# Calculations
sum(all_steps)  # Total steps
mean(all_steps)  # Average daily steps
max(all_steps)  # Maximum steps in a day
all_steps[all_steps > 9000]  # Days with over 9000 steps
```







