---
title: "__Topic 7 - Data Frames in R__"
author: "Esam Mahdi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

# Introduction to Data Frames

In data analysis, information is typically organized in **tables** with rows and columns - just like an Excel spreadsheet or a database table. **R data frames** are specifically designed to hold this kind of tabular data.

## What is a Data Frame?

A **data frame** is a two-dimensional data structure where:

  - **Columns** represent **variables** (features, attributes)

  - **Rows** represent **observations** (records, cases)

  - Different columns can hold **different data types** (numeric, character, logical, etc.)

##Data Frames vs. Other Structures

| Structure | Dimensions | Data Type Restriction | Best Used For |
|-----------|------------|---------------------|----------------|
| Vector | 1D | All elements same type | Single variable data |
| Matrix | 2D | All elements same type | Homogeneous numerical data |
| Data Frame | 2D | Different types per column | Real-world datasets |
| List | 1D | Any type (can be mixed) | Grouping diverse objects |

## How Data Frames Work

Under the hood, a data frame is actually a **list** where:

  - Each **list element** is a **column** (vector)

  - All column vectors must have the **same length**

  - Column vectors can be of **different types**

Think of a data frame as a **spreadsheet** where:

  - Each column is a vector (all same type)

  - Different columns can have different types

  - Each row represents one complete observation

```
Data Frame Example:
     Name     Age  Grade  Passed
1    Alice    22     85    TRUE
2    Bob      19     92    TRUE
3  Charlie    21     67   FALSE

Columns: character, numeric, numeric, logical
Rows: 3 observations
```

# Creating Data Frames

## Basic Creation with `data.frame()`

Pass vectors of the **same length** to the `data.frame()` function. Each vector becomes a column:

```{r}
# Create vectors of the same length
names <- c("Alice", "Bob", "Charlie", "Diana", "Eve")
ages <- c(22, 19, 21, 23, 20)
scores <- c(85, 92, 78, 88, 95)
passed <- c(TRUE, TRUE, FALSE, TRUE, TRUE)

# Create a data frame
students <- data.frame(names, ages, scores, passed)
students
```

**Notice:** The column names are taken from the vector names we used.

## Naming Columns

You can specify column names directly when creating the data frame:

```{r}
# Create data frame with custom column names
students <- data.frame(
  Student_Name = names,
  Age = ages,
  Exam_Score = scores,
  Passed_Exam = passed
)
students
```

## Setting Row Names

You can also give names to the rows:

```{r}
students <- data.frame(
  Student_Name = names,
  Age = ages,
  Exam_Score = scores,
  Passed_Exam = passed,
  row.names = c("S1", "S2", "S3", "S4", "S5")
)
students
```

## Viewing Column and Row Names

```{r}
# Get column names
colnames(students)

# Change column names
colnames(students) <- c("Name", "Age", "Score", "Passed")
students

# Get row names
rownames(students)

# Change row names
rownames(students) <- c("Student1", "Student2", "Student3", "Student4", "Student5")
students
```

## Important Note About Character Columns

By default, R converts character vectors into factors (a special data type for categorical data). To prevent this, use `stringsAsFactors = FALSE`:

```{r}
# Default behavior (converts to factor)
students_default <- data.frame(Name = names, Age = ages)
typeof(students_default$Name)  # Returns "integer" (factor is stored as integer)

# Prevent conversion to factor
students_correct <- data.frame(Name = names, Age = ages, stringsAsFactors = FALSE)
typeof(students_correct$Name)  # Returns "character"

# We'll learn more about factors in the next lesson
```

## Exploring Data Frames with Built-in Datasets

**R** comes with several built-in datasets for practice. We'll use `mtcars` (Motor Trend Car Road Tests) which contains data about 32 cars from the 1970s.

```{r}
# Load the mtcars dataset
cars <- mtcars

# View the first few rows
head(cars)
```

**What's in mtcars?**

  - **mpg**: Miles per gallon (fuel efficiency)

  - **cyl**: Number of cylinders

  - **disp**: Displacement (engine size)

  - **hp**: Horsepower

  - **wt**: Weight (in 1000 lbs)

  - And several other variables...

## Basic Information Functions

```{r}
# Get dimensions (rows, columns)
dim(cars)       # 32 rows, 11 columns

# Number of rows
nrow(cars)      # 32

# Number of columns
ncol(cars)      # 11

# Column names
colnames(cars)

# First few rows (default 6)
head(cars)

# Last few rows
tail(cars)

# View a specific number of rows
head(cars, 3)
tail(cars, 2)
```

## Understanding Data Structure with `str()`

The `str()` function gives a compact summary of the data frame's structure:

```{r}
str(cars)
```

**What `str()` tells us:**

  - `data.frame: 32 obs. of 11 variables` - 32 rows, 11 columns

  - `$ mpg : num` - mpg column is numeric

  - `$ cyl : num` - cyl column is numeric

    And so on...

## Summary Statistics with `summary()`

The `summary()` function provides statistical summaries for each column:

```{r}
summary(cars)
```

**For numeric columns:** `min, 1st quartile, median, mean, 3rd quartile, max`

**For character/factor columns:** counts of each unique value

# Accessing Data Frame Elements

Because data frames are lists under the hood, they support list-style indexing (using `$`, `[[]]`, and `[]`).

## List-Style Indexing

```{r}
# Using $ (returns a vector)
cars$mpg

# Using [[ ]] with column name (returns a vector)
cars[["mpg"]]

# Using [[ ]] with column number (returns a vector)
cars[[1]]

# Using [ ] with column number (returns a data frame)
cars[1]          # Single column as data frame
cars[1:3]        # Multiple columns as data frame
```

**Key Difference:**

  - `[]` returns a data frame (even with one column)

  - `[[]]` or `$` returns the actual vector

```{r}
# Compare the results
result1 <- cars[1]      # Data frame
result2 <- cars[[1]]    # Numeric vector

class(result1)  # "data.frame"
class(result2)  # "numeric"
```

## Matrix-Style Indexing (Rows and Columns)

Use square brackets with a comma separating rows and columns: `[rows, columns]`

```{r}
# Single element: row 5, column 3
cars[5, 3]

# Entire row 5 (all columns)
cars[5, ]

# Entire column 3 (all rows)
cars[, 3]

# Multiple rows: rows 1-5, all columns
cars[1:5, ]

# Multiple columns: all rows, columns 1, 3, 5
cars[, c(1, 3, 5)]

# Range of rows and columns: rows 1-5, columns 2-4
cars[1:5, 2:4]

# Using column names
cars[1:5, c("mpg", "hp", "wt")]

# Using row names (if they exist)
cars["Mazda RX4", ]
```

## Excluding Rows or Columns

Use the minus sign `-` to exclude:

```{r}
# Exclude row 1
cars[-1, ]

# Exclude columns 2 and 4
cars[, -c(2, 4)]

# Exclude first 5 rows
cars[-(1:5), ]

# Keep all rows except first 3, keep all columns except last 2
cars[-(1:3), -(9:11)]
```

## Logical Indexing (Filtering)

Create a logical condition to filter rows:

```{r}
# Find cars with mpg > 25 (good fuel efficiency)
cars[cars$mpg > 25, ]

# Find cars with horsepower > 200
cars[cars$hp > 200, ]

# Combine conditions with & (AND)
cars[cars$mpg > 20 & cars$hp > 100, ]

# Combine conditions with | (OR)
cars[cars$cyl == 4 | cars$cyl == 8, ]

# Find cars with weight less than 2.5 tons
cars[cars$wt < 2.5, ]
```

## The `subset()` Function

The `subset()` function provides an alternative way to filter:

```{r}
# Basic filtering
subset(cars, mpg > 25)

# Multiple conditions
subset(cars, mpg > 20 & hp > 100)

# Select specific columns as well
subset(cars, mpg > 20, select = c(mpg, hp, wt))

# Exclude columns
subset(cars, mpg > 20, select = -c(vs, am, gear, carb))
```

# Modifying Data Frames

## Adding New Columns

Use `$` or `[[]]` to add a new column:

```{r}
# Create a fresh copy of mtcars
cars <- mtcars

# Add horsepower per cylinder column
cars$hp_per_cylinder <- cars$hp / cars$cyl

# Add power-to-weight ratio
cars$power_to_weight <- cars$hp / cars$wt

# View the new columns
head(cars[, c("hp", "cyl", "hp_per_cylinder", "power_to_weight")])
```

## Adding Columns with Matrix Indexing

```{r}
# Add a column using matrix indexing
cars[, "mpg_squared"] <- cars$mpg^2

# Add a logical column
cars$is_efficient <- cars$mpg > 25

head(cars[, c("mpg", "mpg_squared", "is_efficient")])
```

## Removing Columns

Set a column to `NULL` to remove it:

```{r}
# Remove single column
cars$mpg_squared <- NULL

# Remove multiple columns
cars$hp_per_cylinder <- NULL
cars$power_to_weight <- NULL

head(cars)
```

## Removing Rows

Use matrix indexing with negative indices or logical conditions:

```{r}
# Create a fresh copy
cars <- mtcars

# Remove row 1 (Mazda RX4)
cars_without_first <- cars[-1, ]
head(cars_without_first, 3)

# Remove rows 1 through 5
cars_without_first5 <- cars[-(1:5), ]
head(cars_without_first5, 3)

# Remove cars with low mpg (less than 15)
cars_high_mpg <- cars[cars$mpg >= 15, ]
nrow(cars_high_mpg)  # Count of remaining cars
```

## Combining Data Frames

**`rbind()` - Stack rows vertically**

```{r}
# Create two data frames with same columns
df1 <- data.frame(ID = 1:3, Name = c("A", "B", "C"))
df2 <- data.frame(ID = 4:6, Name = c("D", "E", "F"))

# Combine by rows
combined_rows <- rbind(df1, df2)
combined_rows
```

**`cbind()` - Combine columns horizontally**

```{r}
# Create two data frames with same number of rows
df_a <- data.frame(ID = 1:4, Score = c(85, 90, 78, 92))
df_b <- data.frame(Grade = c("B", "A", "C", "A"))

# Combine by columns
combined_cols <- cbind(df_a, df_b)
combined_cols
```

## Common Data Frame Operations

**Summary Functions on Columns**

```{r}
# Calculate statistics on numeric columns
mean(cars$mpg)        # Average mpg
median(cars$hp)       # Median horsepower
sd(cars$wt)           # Standard deviation of weight
min(cars$disp)        # Minimum displacement
max(cars$qsec)        # Maximum quarter mile time

# Multiple statistics
summary(cars$mpg)

# Column-wise operations
colMeans(cars[, c("mpg", "hp", "wt")])
colSums(cars[, c("mpg", "hp", "wt")])
```

**Adding Meaningful Row Names**

```{r}
# Create a data frame with student data
students <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Score = c(92, 87, 95),
  Grade = c("A", "B", "A"),
  stringsAsFactors = FALSE
)

# Add meaningful row names (e.g., student IDs)
rownames(students) <- c("STU001", "STU002", "STU003")
students

# Access by row name
students["STU002", ]
```

**Sorting Data Frames**
  
```{r}
# Sort by mpg (ascending order)
cars_sorted_mpg <- cars[order(cars$mpg), ]
head(cars_sorted_mpg[, c("mpg")])

# Sort by mpg (descending order)
cars_sorted_mpg_desc <- cars[order(-cars$mpg), ]
head(cars_sorted_mpg_desc[, c("mpg")])

# Sort by multiple columns (first by cyl, then by mpg)
cars_sorted <- cars[order(cars$cyl, cars$mpg), ]
head(cars_sorted[, c("cyl", "mpg")])
```

# Examples

## Example 1: Student Grade Analysis

```{r}
# Create student dataset
students <- data.frame(
  Student_ID = 1001:1008,
  Name = c("Emma", "Liam", "Sophia", "Noah", "Olivia", "Mason", "Ava", "Ethan"),
  Math = c(85, 92, 78, 88, 95, 82, 91, 87),
  Science = c(88, 85, 92, 79, 94, 88, 86, 90),
  English = c(82, 90, 85, 91, 88, 84, 92, 89),
  Attendance = c(95, 98, 92, 96, 99, 94, 97, 93),
  stringsAsFactors = FALSE
)

# Add average score column
students$Average <- (students$Math + students$Science + students$English) / 3

# Add letter grade column
students$Letter_Grade <- ifelse(students$Average >= 90, "A",
                         ifelse(students$Average >= 80, "B",
                         ifelse(students$Average >= 70, "C", "F")))

# Find top 3 students
top_students <- students[order(-students$Average), ][1:3, ]
top_students[, c("Name", "Average", "Letter_Grade")]

# Find students with high attendance (>95%)
high_attendance <- students[students$Attendance > 95, ]
high_attendance[, c("Name", "Attendance", "Average")]

# Summary statistics
summary(students[, c("Math", "Science", "English", "Average", "Attendance")])
```

## Example 2: Sales Data Analysis

```{r}
# Create sales dataset
sales <- data.frame(
  Date = c("Jan", "Jan", "Jan", "Feb", "Feb", "Feb", "Mar", "Mar", "Mar"),
  Product = c("A", "B", "C", "A", "B", "C", "A", "B", "C"),
  Units_Sold = c(150, 200, 120, 180, 220, 140, 200, 250, 160),
  Price = c(10, 15, 20, 10, 15, 20, 10, 15, 20),
  stringsAsFactors = FALSE
)

# Calculate Revenue
sales$Revenue <- sales$Units_Sold * sales$Price

# View the data
sales

# Total revenue by product
aggregate(Revenue ~ Product, data = sales, sum)

# Average units sold by month
aggregate(Units_Sold ~ Date, data = sales, mean)

# Best selling month for each product
for(product in unique(sales$Product)) {
  product_data <- sales[sales$Product == product, ]
  best_month <- product_data[which.max(product_data$Units_Sold), "Date"]
  cat("Product", product, "sold best in", best_month, "\n")
}
```

# Common Mistakes to Avoid

## Mistake 1: Confusing `[]` and `[[]]`

```{r}
df <- data.frame(x = 1:3, y = 4:6)

# Wrong: returns a data frame, not a vector
result_wrong <- df[1]
result_wrong + 10  # Error!

# Correct: returns the actual vector
result_correct <- df[[1]]
result_correct + 10  # Works: 11 12 13
```

## Mistake 2: Forgetting the Comma in Matrix Indexing

```{r}
df <- data.frame(A = 1:3, B = 4:6, C = 7:9)

# Wrong: missing comma means select columns (returns data frame)
df[1:2]  # Returns columns 1-2

# Correct: comma means rows 1-2, all columns
df[1:2, ]  # Returns rows 1-2
```

## Mistake 3: Mixing Data Types in a Column (Impossible!)

```{r}
# This will coerce all elements to the same type
mixed <- data.frame(
  x = c(1, "two", 3),  # All become character
  stringsAsFactors = FALSE
)
mixed$x  # "1", "two", "3" (all text)
typeof(mixed$x)  # "character"
```

## Mistake 4: Using `$` with Computed Names

```{r}
df <- data.frame(score = 1:5, grade = c("F", "D", "C", "B", "A"))

column_name <- "score"

# Wrong: looks for column literally named 'column_name'
df$column_name  # NULL

# Correct: use [[]] for variable column names
df[[column_name]]  # Returns the score column
```

# Summary Table

| Operation | Syntax | Example | Result |
|-----------|--------|---------|--------|
| Create data frame | `data.frame()` | `data.frame(x = 1:3, y = 4:6)` | New data frame |
| Prevent factor conversion | `stringsAsFactors = FALSE` | `data.frame(x = "a", stringsAsFactors = FALSE)` | Character column |
| View first rows | `head()` | `head(df, 10)` | First 10 rows |
| View structure | `str()` | `str(df)` | Compact summary |
| Summary statistics | `summary()` | `summary(df)` | Statistical summary |
| Dimensions | `dim()` | `dim(df)` | Rows and columns |
| Column names | `colnames()` | `colnames(df)` | Vector of column names |
| Row names | `rownames()` | `rownames(df)` | Vector of row names |
| Access column (vector) | `$` or `[[ ]]` | `df$col` or `df[["col"]]` | Vector |
| Access column (data frame) | `[ ]` | `df["col"]` | Data frame |
| Access element | `[row, col]` | `df[2, 3]` | Single value |
| Access row | `[row, ]` | `df[2, ]` | Data frame row |
| Access column | `[, col]` | `df[, 3]` | Vector |
| Subset rows | `[condition, ]` | `df[df$x > 5, ]` | Filtered data frame |
| Subset with function | `subset()` | `subset(df, x > 5)` | Filtered data frame |
| Add column | `$new <-` | `df$new <- 1:5` | Modified data frame |
| Remove column | `<- NULL` | `df$col <- NULL` | Modified data frame |
| Remove rows | `[-rows, ]` | `df[-1, ]` | Data frame without row 1 |
| Sort | `order()` | `df[order(df$x), ]` | Sorted data frame |
| Combine rows | `rbind()` | `rbind(df1, df2)` | Stacked data frame |
| Combine columns | `cbind()` | `cbind(df1, df2)` | Joined data frame |

## Key Summaries

  - **Data frames** are R's version of spreadsheets/tables

  - **Columns** can have different data types (numeric, character, logical)

  - Data frames are **lists** under the hood (each column is a list element)

  - Use `data.frame()` to create data frames manually

  - `str()` and `summary()` help explore new datasets

  - `$` and `[[]]` access columns as vectors

  - `[rows, columns]` provides matrix-style indexing

  - **Logical indexing** is powerful for filtering rows

  - `rbind()` and `cbind()` combine data frames

  - `stringsAsFactors = FALSE` prevents automatic conversion to factors

# Exercises
## Exercise 1: Create and Explore a Data Frame

Create a data frame called employees with the following columns:

  - `name`: "John", "Sarah", "Mike", "Lisa", "David"

  - `age`: 28, 34, 42, 29, 38

  - `department`: "Sales", "IT", "HR", "Marketing", "IT"

  - `salary`: 55000, 75000, 65000, 58000, 82000

  - `full_time`: TRUE, TRUE, FALSE, TRUE, TRUE

Then:

  - View the structure of the data frame

  - View a summary of the data frame

  - Display the first 3 rows

```{r}
# Your code here:
```

## Solution 

```{r}
employees <- data.frame(
  name = c("John", "Sarah", "Mike", "Lisa", "David"),
  age = c(28, 34, 42, 29, 38),
  department = c("Sales", "IT", "HR", "Marketing", "IT"),
  salary = c(55000, 75000, 65000, 58000, 82000),
  full_time = c(TRUE, TRUE, FALSE, TRUE, TRUE),
  stringsAsFactors = FALSE
)

str(employees)
summary(employees)
head(employees, 3)
```

## Exercise 2: Indexing and Subsetting

Using the `employees` data frame from Exercise 1:

  - Extract the entire `salary` column as a vector

  - Extract the `department` column as a data frame

  - Extract the row for "Mike"

  - Extract the names and salaries of employees over 35

  - Extract employees in the IT department who are over 35

```{r}
# Your code here:
```

## Solution 

```{r}
# Salary as vector
employees$salary

# Department as data frame
employees["department"]

# Row for Mike
employees[employees$name == "Mike", ]

# Names and salaries of employees over 35
employees[employees$age > 35, c("name", "salary")]

# IT employees over 35
employees[employees$department == "IT" & employees$age > 35, ]
```

## Exercise 3: Modifying Data Frames

Start with the `employees` data frame. Then:

  - Add a new column called `bonus` equal to 10% of salary

  - Add a column called `total_compensation` equal to salary + bonus

  - Update the salary of "Sarah" to 78000

  - Remove the `full_time` column

  - Add a new row for "Anna" (age 31, department "Sales", salary 62000)

```{r}
# Your code here:
```

## Solution 

```{r}
# Add bonus and total_compensation
employees$bonus <- employees$salary * 0.10
employees$total_compensation <- employees$salary + employees$bonus

# Update Sarah's salary
employees[employees$name == "Sarah", "salary"] <- 78000
employees[employees$name == "Sarah", "total_compensation"] <- employees[employees$name == "Sarah", "salary"] * 1.10

# Remove full_time column
employees$full_time <- NULL

# Add new row for Anna
new_row <- data.frame(
  name = "Anna",
  age = 31,
  department = "Sales",
  salary = 62000,
  bonus = 6200,
  total_compensation = 68200,
  stringsAsFactors = FALSE
)
employees <- rbind(employees, new_row)
employees
```

## Exercise 4: Working with mtcars

Using the built-in `mtcars` dataset:

  - Find all cars with 4 cylinders and mpg greater than 25

  - Find all cars with horsepower between 100 and 150

  - Add a new column called `displacement_per_cylinder` (disp divided by cyl)

  - Find the car with the highest horsepower per cylinder

  - Calculate the average mpg for cars with automatic (am = 0) vs manual (am = 1) transmission

```{r}
# Your code here:
```

## Solution

```{r}
# 4-cylinder cars with mpg > 25
mtcars[mtcars$cyl == 4 & mtcars$mpg > 25, ]

# Cars with hp between 100 and 150
mtcars[mtcars$hp >= 100 & mtcars$hp <= 150, ]

# Add displacement per cylinder
mtcars$disp_per_cyl <- mtcars$disp / mtcars$cyl

# Car with highest hp per cylinder
mtcars$hp_per_cyl <- mtcars$hp / mtcars$cyl
mtcars[which.max(mtcars$hp_per_cyl), ]

# Average mpg by transmission type
mtcars_auto <- mtcars[mtcars$am == 0, "mpg"]
mtcars_manual <- mtcars[mtcars$am == 1, "mpg"]
mean(mtcars_auto)
mean(mtcars_manual)
```

## Exercise 5: Combining Data Frames

Create two data frames:

  - `sales_Q1` with columns: `product` ("A", "B", "C"), `revenue` (100, 150, 120)

  - `sales_Q2` with columns: `product` ("A", "B", "C"), `revenue` (110, 160, 130)

Then:

  - Combine them vertically to show all sales

  - Add a new column called `quarter` to each before combining

  - Combine them horizontally to compare Q1 and Q2 side by side

```{r}
# Your code here:
```

## Solution

```{r}
# Create sales data frames
sales_Q1 <- data.frame(product = c("A", "B", "C"), revenue = c(100, 150, 120))
sales_Q2 <- data.frame(product = c("A", "B", "C"), revenue = c(110, 160, 130))

# Combine vertically
all_sales <- rbind(sales_Q1, sales_Q2)
all_sales

# Add quarter column
sales_Q1$quarter <- "Q1"
sales_Q2$quarter <- "Q2"
all_sales_with_q <- rbind(sales_Q1, sales_Q2)
all_sales_with_q

# Combine horizontally
colnames(sales_Q2)[2] <- "revenue_Q2"
combined <- cbind(sales_Q1, sales_Q2[, "revenue_Q2", drop = FALSE])
combined
```

## Exercise 6: Analysis Challenge

Using the `mtcars` dataset, answer these questions:

  - What is the average horsepower of cars with 8 cylinders? With 4 cylinders?

  - Which car has the best fuel efficiency (highest mpg)?

  - Create a subset of cars with weight less than 3 tons (3000 lbs)

  - Which cylinder group (4, 6, or 8) has the highest average weight?

  - Find all cars that are both powerful (hp > 150) and heavy (wt > 4)

```{r}
# Your code here:
```

## Solution

```{r}
# Average horsepower by cylinder
mean(mtcars[mtcars$cyl == 8, "hp"])
mean(mtcars[mtcars$cyl == 4, "hp"])

# Best fuel efficiency
mtcars[which.max(mtcars$mpg), ]

# Cars under 3 tons (3000 lbs = 3.0 in wt since wt is in 1000 lbs)
light_cars <- mtcars[mtcars$wt < 3, ]
light_cars

# Average weight by cylinder group
aggregate(wt ~ cyl, data = mtcars, mean)

# Powerful and heavy cars
mtcars[mtcars$hp > 150 & mtcars$wt > 4, ]
```



