---
title: "__Topic 6: Lists in R__"
author: "Esam Mahdi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

# Introduction to Lists

In previous lectures, we learned about **vectors** (one-dimensional, same type) and **matrices/arrays** (multi-dimensional, same type). But real-world data often requires storing **different types of data together:**

  - A student record: name (character), age (numeric), grades (numeric vector), and photo (matrix)

  - A product inventory: product ID (numeric), name (character), price (numeric), in_stock (logical)

  - A model output: coefficients (numeric vector), residuals (numeric vector), formula (character), and R-squared (numeric)

**Lists** are R's solution to this problem. A list is a data structure that can contain **any combination of objects of different types and sizes** - numbers, text, vectors, matrices, arrays, or even other lists!

## Lists vs. Other Data Structures

| Structure	| Homogeneous? | Dimensions | Can contain lists? |
|-----------|--------------|-----------------|---------|
| Vector | Yes (all same type)	| 1D | No |
| Matrix/Array | Yes (all same type) | 2D or more	| No |
| **List** | **No (can mix types)**	| **1D (but each element can be complex)** | **Yes** |

Think of a list as a **filing cabinet** where each drawer can hold something different:

  - Drawer 1: A single number

  - Drawer 2: A whole vector of test scores

  - Drawer 3: A matrix of grades

  - Drawer 4: Another list of student details

# Creating Lists

## Basic List Creation with `list()`

Use the `list()` function and pass each object you want to store, separated by commas:

```{r}
# Create a list with different types of objects
my_list <- list(
  1:5,                          # Numeric vector
  "Hello, R!",                  # Character string
  matrix(1:6, nrow = 2, ncol = 3)  # Matrix
)
# Display the list
my_list
```

**Understanding the output:**

  - [[1]] indicates the first element of the list

  - [1] 1 2 3 4 5 is the content (a numeric vector)

  - [[2]] is the second element (a character string)

  - [[3]] is the third element (a 2×3 matrix)

## Creating Named Lists

You can give names to each element when creating the list. This makes accessing elements much easier:

```{r}
# Create a named list
student_info <- list(
  name = "Alice Johnson",
  age = 22,
  grades = c(85, 90, 88, 92),
  is_enrolled = TRUE,
  courses = matrix(c("Math", "Physics", "Chemistry"), nrow = 3, ncol = 1)
)

student_info
```

## Creating Empty Lists

Sometimes you want to create an empty list and fill it later:

```{r}
# Create an empty list
empty_list <- list()
empty_list

# Create a list with a specific length (all elements NULL)
preallocated <- vector("list", length = 5)
preallocated
```

## Understanding List Structure

**The `str()` Function - See List Structure**

The `str()` function (structure) gives you a compact, readable summary of a list:

```{r}
str(student_info)
```

What `str()` tells you:

  - `List of 5` - the list has 5 elements

  - `$ name: chr "Alice Johnson"` - element named "name" is character

  - `$ age: num 22` - element named "age" is numeric

  - `$ grades: num [1:4] 85 90 88 92` - numeric vector of length 4

  - `$ is_enrolled: logi TRUE` - logical value

  - `$ courses: chr [1:3, 1] "Math" "Physics" "Chemistry"` - character matrix

**The `summary()` Function**

```{r}
summary(student_info)
```

**The `length()` Function**

```{r}
# Number of elements in the list (not total items inside elements)
length(student_info)  # Returns 5 (name, age, grades, is_enrolled, courses)

# Compare with length of a vector inside the list
length(student_info$grades)  # Returns 4
```

## Accessing List Elements

There are **three ways** to access elements in a list, each with different behavior.

**Method 1: Single Bracket `[]` - Returns a Sublist**

Using single brackets returns a **new list** containing the selected elements:

```{r}
# Create a sample list
my_list <- list(
  first = c(10, 20, 30),
  second = "Important text",
  third = matrix(1:4, nrow = 2)
)
my_list

# Extract first two elements as a list
subset_list <- my_list[1:2]
subset_list
typeof(subset_list)  # Still a list

# Extract single element as a list
single_element_list <- my_list[2]
single_element_list
typeof(single_element_list)  # Still a list (not the actual string!)
```

**Key point:** `my_list[2]` gives you a list containing the second element, NOT the actual value.

**Method 2: Double Bracket `[[]]` - Returns the Actual Object**

Double brackets extract the **actual object** inside the list element:

```{r}
# Extract the second element's actual value
actual_value <- my_list[[2]]
actual_value
typeof(actual_value)  # Character, not list!

# Extract the third element (matrix)
my_list[[3]]
```

**Method 3: Dollar Sign `$` - Access by Name (Recommended)**

If your list has names, `$` is the most readable way to access elements:

```{r}
# Access by name
my_list$first
my_list$second
my_list$third

# The result is the actual object, not a sublist
typeof(my_list$first)   # "double" (numeric vector)
typeof(my_list$second)  # "character"
```

## Comparing Access Methods

```{r}
# Create a named list
person <- list(
  name = "Bob",
  scores = c(95, 87, 92)
)

# Three different ways to access the scores:

# Method 1: Single bracket (returns list)
person[2]        # Returns a list containing the scores
class(person[2]) # "list"

# Method 2: Double bracket (returns vector)
person[[2]]      # Returns the actual scores vector
class(person[[2]]) # "numeric"

# Method 3: Dollar sign (returns vector)
person$scores    # Returns the actual scores vector
class(person$scores) # "numeric"

# When to use each:
# - Use $ or [[ ]] when you want to work with the actual data
# - Use [ ] when you want to extract a sublist (e.g., for combining lists)
```

## Accessing Elements Inside Elements

Once you extract an element with `[[]]` or `$`, you can immediately index into it:

```{r}
# Access the third element, then get row 2, column 2
my_list[[3]][2, 2]

# Access the first element, then get the second value
my_list$first[2]

# Access the grades vector, then get the third grade
student_info$grades[3]

# Access the matrix in the third element, then get the first row
my_list[[3]][1, ]
```

## Accessing Non-Existent Elements

```{r}
# Returns NULL (not an error)
my_list$nonexistent

# Returns NULL with a warning
my_list[["nonexistent"]]
```

```
# Trying to access beyond list length returns NULL
my_list[[10]]  # NULL (no error)
```

## Modifying Lists

### Adding New Elements

You can add new elements using `$`, `[[]]`, or `[]`:

```{r}
# Start with an empty list
my_list <- list()
my_list

# Add elements using $
my_list$name <- "Data Science Course"
my_list$students <- c("Alice", "Bob", "Charlie")
my_list

# Add using double brackets
my_list[[4]] <- c(95, 87, 92)  # No name
my_list

# Add using single brackets (must use a list on the right)
my_list[5] <- list("Additional info")
my_list

# Add with a new name
my_list$instructor <- "Dr. Smith"
```

### Removing Elements
Set an element to `NULL` to remove it:

```{r}
# Remove by name
my_list$name <- NULL
my_list

# Remove by position
my_list[[2]] <- NULL
my_list

# Remove multiple elements
my_list[c(1, 3)] <- NULL
my_list
```

### Updating Existing Elements

Just reassign to the same position:

```{r}
# Create a list
scores <- list(
  math = 85,
  science = 90,
  history = 78
)
scores

# Update a value
scores$math <- 92
scores

# Update using position
scores[[2]] <- 95
scores

# Replace entire element with different type
scores$history <- "A-"
scores
```

### Combining Lists with `c()`

The `c()` function (combine) can join two or more lists:

```{r}
# Create two lists
list1 <- list(a = 1, b = 2, c = 3)
list2 <- list(d = 4, e = 5, f = 6)

# Combine them
combined <- c(list1, list2)
combined
str(combined)

# Combine list with non-list elements
combined2 <- c(list1, x = 100, y = 200)
combined2
```

## Working with List Names

### Getting and Setting Names

```{r}
# Create a list without names
unnamed_list <- list(10, "text", matrix(1:4, nrow = 2))
unnamed_list

# Check names (initially NULL)
names(unnamed_list)

# Assign names
names(unnamed_list) <- c("number", "words", "my_matrix")
unnamed_list

# Access names
names(unnamed_list)

# Change a specific name
names(unnamed_list)[2] <- "phrase"
unnamed_list
```

### Nested Lists (Lists Within Lists)

Lists can contain other lists - this is called **nesting:**

```{r}
# Create a nested list
course <- list(
  course_name = "R Programming",
  instructor = list(
    name = "Dr. Jones",
    email = "jones@university.edu",
    office = "Room 301"
  ),
  students = list(
    student1 = list(name = "Alice", grade = 92),
    student2 = list(name = "Bob", grade = 87),
    student3 = list(name = "Charlie", grade = 95)
  ),
  schedule = c("Monday", "Wednesday", "Friday")
)

# Examine structure
str(course)

# Access nested elements
course$instructor$name
course$students$student2$grade

# Access using multiple brackets
course[[2]][[2]]  # email
course[["students"]][["student3"]][["name"]]
```

## Common List Functions

**`lapply()` - Apply Function to Each List Element**

`lapply()` applies a function to every element of a list and returns a list:

```{r}
# Create a list of numeric vectors
numbers <- list(
  a = c(1, 2, 3),
  b = c(10, 20, 30, 40),
  c = c(100, 200)
)
numbers

# Calculate mean of each vector
lapply(numbers, mean)

# Calculate sum of each vector
lapply(numbers, sum)

# Get length of each vector
lapply(numbers, length)

# Apply a custom function
lapply(numbers, function(x) x * 2)
```

**`sapply()` - Simplified Version of `lapply()`**

`sapply()` tries to simplify the result to a vector or matrix if possible:

```{r}
# sapply returns a vector (simpler)
sapply(numbers, mean)

# Compare lapply (returns list) vs sapply (returns vector)
lapply(numbers, mean)  # List
sapply(numbers, mean)  # Named numeric vector

# When results have same length, sapply creates a matrix
sapply(numbers, function(x) c(mean(x), sd(x)))
```

**`unlist()` - Convert List to Vector**

```{r}
# Nested list
nested <- list(a = 1, b = list(c = 2, d = 3), e = 4)
nested

# Unlist to a single vector
unlist(nested)

# With character elements
mixed <- list(name = "John", age = 25, city = "New York")
unlist(mixed)  # All become character
```

# Examples

## Example 1: Student Database

```{r}
# Create a list of student records
student_database <- list(
  student1 = list(
    name = "Emma Watson",
    id = 1001,
    major = "Computer Science",
    gpa = 3.8,
    courses_taken = c("CS101", "CS201", "MATH150")
  ),
  student2 = list(
    name = "Liam Chen",
    id = 1002,
    major = "Statistics",
    gpa = 3.9,
    courses_taken = c("STAT200", "MATH150", "CS101")
  ),
  student3 = list(
    name = "Sophia Rodriguez",
    id = 1003,
    major = "Data Science",
    gpa = 3.7,
    courses_taken = c("DS101", "CS201", "STAT200")
  )
)

# View structure
str(student_database, max.level = 1)

# Access specific information
student_database$student2$name
student_database$student1$courses_taken

# Get all student names
sapply(student_database, function(x) x$name)

# Get all GPAs
sapply(student_database, function(x) x$gpa)

# Find students with GPA > 3.8
high_gpa <- sapply(student_database, function(x) x$gpa > 3.8)
names(student_database[high_gpa])
```

## Example 2: Statistical Model Output

Many **R** functions return lists. Here's an example of a simple regression output:

```{r}

# Create sample data
x <- 1:10
y <- 2 * x + 3 + rnorm(10, sd = 1)

# Simple linear regression (lm() returns a list)
model <- lm(y ~ x)

# Check what lm() returns
str(model, max.level = 1)

# Access different components
model$coefficients
model$residuals
model$fitted.values
summary(model)$r.squared

# Create a custom summary list
analysis_results <- list(
  coefficients = coef(model),
  r_squared = summary(model)$r.squared,
  p_value = summary(model)$coefficients[2, 4],
  residuals = residuals(model),
  sample_size = length(x)
)

analysis_results
```

## Example 3: Configuration Settings

```{r}
# Use lists for configuration
config <- list(
  database = list(
    host = "localhost",
    port = 5432,
    name = "sales_db",
    user = "admin"
  ),
  analysis = list(
    alpha = 0.05,
    n_bootstrap = 1000,
    random_seed = 42
  ),
  paths = list(
    input_dir = "/data/input",
    output_dir = "/data/output",
    logs_dir = "/data/logs"
  ),
  flags = list(
    verbose = TRUE,
    save_plots = TRUE,
    parallel = FALSE
  )
)

# Access configuration values
config$database$host
config$analysis$alpha
config$paths$output_dir

# Modify config
config$flags$parallel <- TRUE
config$analysis$n_bootstrap <- 5000
```

# Common Mistakes

## Mistake 1: Confusing `[` and `[[`

```{r}
my_list <- list(a = 1:3, b = "hello")

# Wrong when you need the actual value
result1 <- my_list[1]   # Returns a list
```

```
result1 + 5  # Error: non-numeric argument
```

```{r}
# Correct when you need the value
result2 <- my_list[[1]] # Returns numeric vector
result2 + 5  # Works: 6 7 8

# Correct with names
result3 <- my_list$a   # Returns numeric vector
result3 + 5  # Works
```

## Mistake 2: Using $ with Computed Names

```{r}
my_list <- list(first = 10, second = 20)

# This works (literal name)
my_list$first

# This doesn't work as expected (dollar doesn't evaluate variables)
name <- "first"
my_list$name  # Looks for element named "name", not "first"

# Use double brackets for computed names
my_list[[name]]  # Works: returns 10
```

## Mistake 3: Forgetting Lists Can Hold Anything

```{r}
# This is valid but can be confusing
crazy_list <- list(
  number = 42,
  vector = 1:5,
  matrix = matrix(1:4, nrow = 2),
  list = list(a = 1, b = 2),
  function_example = function(x) x^2
)

crazy_list$function_example(10)  # Call the function: returns 100
```

## Mistake 4: Assuming unlist() Preserves Types

```{r}
mixed <- list(age = 25, name = "John", active = TRUE)
unlist(mixed)  # All become character!
```

**When to Use Lists**

Use lists when:

  - You need to group different types of data (e.g., student records)

  - Functions return multiple outputs (like lm(), eigen(), svd())

  - You need flexible, hierarchical data structures

  - You're building configuration objects

  - You need to store models or analysis results

**Don't use lists when:**

  - You only need one type of data (use vectors or matrices instead)

  - Performance is critical (lists have more overhead)

  - You need to do vectorized calculations (use vectors/matrices)

# Summary Table

| Operation | Syntax | Example | Result |
|-----------|--------|---------|--------|
| Create list | `list()` | `list(1, "a", matrix)` | New list |
| Named list | `list(name = value)` | `list(x = 1:5, y = "text")` | Named list |
| Access element (sublist) | `[ ]` | `my_list[1]` | List with element |
| Access element (value) | `[[ ]]` | `my_list[[1]]` | Actual object |
| Access by name (value) | `$` | `my_list$name` | Actual object |
| Add element | `$` or `[[ ]]` | `my_list$new <- 10` | Modified list |
| Remove element | `<- NULL` | `my_list$old <- NULL` | Modified list |
| List length | `length()` | `length(my_list)` | Number of elements |
| View structure | `str()` | `str(my_list)` | Compact display |
| Get names | `names()` | `names(my_list)` | Character vector |
| Set names | `names() <-` | `names(my_list) <- c("a", "b")` | Named list |
| Combine lists | `c()` | `c(list1, list2)` | New combined list |
| Apply function | `lapply()` | `lapply(my_list, sum)` | List of results |
| Simplified apply | `sapply()` | `sapply(my_list, mean)` | Vector/matrix |
| Convert to vector | `unlist()` | `unlist(my_list)` | Vector |

## Key Summaries

  - Lists are heterogeneous - they can hold different types of data

  - Three ways to access elements: `[]` (sublist), `[[]]` (actual value), `$` (named access)

  - Lists can contain lists - creating nested/hierarchical structures

  - Use `str()` to understand list structure

  - `lapply()` and `sapply()` apply functions to all list elements

  - Many **R** functions return lists (e.g., `lm()`, `eigen()`, `svd()`)

  - Set elements to NULL to remove them from a list

  - `c()` combines lists into a single list

  - Lists are the foundation for data frames (next lecture!)

# Exercises
## Exercise 1: Create a List

Create a list called `my_list` containing:

  - A character string "Learning R"

  - A numeric vector with values 5, 10, 15, 20

  - A 3x3 matrix with numbers 1 through 9

  - A logical value TRUE

Then display the structure of your list.

```{r}
# Your code here:
```

## Solution

```{r}
my_list <- list(
  "Learning R",
  c(5, 10, 15, 20),
  matrix(1:9, nrow = 3, ncol = 3),
  TRUE
)
str(my_list)
```

## Exercise 2: Access List Elements

Using the list from Exercise 1, extract:

  - The second element (as a vector, not as a list)

  - The third element (the matrix)

  - The value at row 2, column 3 of the matrix inside the list

```{r}
# Your code here:
```

## Solution 

```{r}
# Extract second element as vector
my_list[[2]]

# Extract third element (matrix)
my_list[[3]]

# Extract row 2, column 3 from the matrix
my_list[[3]][2, 3]
```

## Exercise 3: Named List

Create a named list called employee with the following elements:

  - `name`: "Jane Smith"

  - `id`: 1045

  - `department`: "Marketing"

  - `salary`: 65000

  - `skills`: a vector containing "R", "Python", "SQL"

  - `full_time`: TRUE

Then:

  - Add a new element `years_experience` with value 5

  - Update the salary to 72000

  - Remove the `full_time` element

```{r}
# Your code here:
```

## Solution 

```{r}
employee <- list(
  name = "Jane Smith",
  id = 1045,
  department = "Marketing",
  salary = 65000,
  skills = c("R", "Python", "SQL"),
  full_time = TRUE
)

# Add years_experience
employee$years_experience <- 5

# Update salary
employee$salary <- 72000

# Remove full_time
employee$full_time <- NULL

employee
```

## Exercise 4: Nested List

Create a nested list called company with the following structure:

  - `name`: "Tech Solutions Inc."

  - `founded`: 2010

  - `departments`: a list containing:

    - `HR`: a list with `manager = "Alice"` and `employees = 5`

    - `IT`: a list with `manager = "Bob"` and `employees = 12`

    - `Sales`: a list with `manager = "Carol"` and `employees = 8`

Then extract:

  - The IT manager's name

  - The number of employees in Sales

  - All department names

```{r}
# Your code here:
```

## Solution 

```{r}
company <- list(
  name = "Tech Solutions Inc.",
  founded = 2010,
  departments = list(
    HR = list(manager = "Alice", employees = 5),
    IT = list(manager = "Bob", employees = 12),
    Sales = list(manager = "Carol", employees = 8)
  )
)

# IT manager's name
company$departments$IT$manager

# Number of employees in Sales
company$departments$Sales$employees

# All department names
names(company$departments)
```

## Exercise 5: Using `lapply()` and `sapply()`

Create a list of three numeric vectors:

    Vector 1: 1 through 10

    Vector 2: 5 through 15

    Vector 3: 100, 200, 300, 400, 500

Then use `lapply()` and `sapply()` to:

    Calculate the mean of each vector

    Calculate the sum of each vector

    Calculate the length of each vector

    Find the maximum value in each vector

Compare the outputs of `lapply()` and `sapply()`.

```{r}
# Your code here:
```

## Solution 

```{r}
vectors_list <- list(
  vec1 = 1:10,
  vec2 = 5:15,
  vec3 = c(100, 200, 300, 400, 500)
)

# Using lapply (returns list)
lapply(vectors_list, mean)
lapply(vectors_list, sum)
lapply(vectors_list, length)
lapply(vectors_list, max)

# Using sapply (returns vector)
sapply(vectors_list, mean)
sapply(vectors_list, sum)
sapply(vectors_list, length)
sapply(vectors_list, max)
```

## Exercise 6: Combining and Unlisting

Create two lists:

  - `list_a` with elements: `x = 1, y = 2, z = 3`

  - `list_b` with elements: `a = "apple", b = "banana", c = "cherry"`

Then:

  - Combine them into one list called `combined`

  - Add a new element `d = 100` to `combined`

  - Remove the `y` element from `combined`

  - Convert `combined` to a vector using `unlist()` (what happens to the data types?)

```{r}
# Your code here:
```

## Solution

```{r}
list_a <- list(x = 1, y = 2, z = 3)
list_b <- list(a = "apple", b = "banana", c = "cherry")

# Combine lists
combined <- c(list_a, list_b)
combined

# Add new element
combined$d <- 100
combined

# Remove y element
combined$y <- NULL
combined

# Convert to vector (everything becomes character!)
unlist(combined)
```


