---
title: "__Topic 2: Atomic Data Types__"
author: "Esam Mahdi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

# The Five Basic Data Types

**R** has five fundamental (atomic) data types. Think of these as the building blocks for all data in **R**:

| Data Type | What It Stores | Example |
|-----------|---------------|---------|
| Numeric | Decimal numbers | `3.14`, `-2.5`, `Inf` |
| Integer | Whole numbers | `42L`, `56L` |
| Character | Text | `"hello"`, `'R is fun'` |
| Logical | True/False values | `TRUE`, `FALSE` |
| Complex | Numbers with imaginary part | `2 + 3i` |

## How to Check Data Types

Use two main functions to identify data types:

    class() - Tells you the high-level data type

    typeof() - Tells you the low-level storage type

```{r}
# Examples with class()
class(3.14)    # Numeric
class(42L)     # Integer
class("hello") # Character
class(TRUE)    # Logical
class(2 + 3i)  # Complex
```

```{r}
# Examples with typeof()
typeof(3.14)   # "double" (numeric)
typeof(42L)    # "integer"
typeof("hello") # "character"
typeof(TRUE)   # "logical"
typeof(2 + 3i) # "complex"
```

**Key Point:** `class()` is more commonly used in data analysis as it gives a more intuitive answer for most **R** objects.

# Numeric (Double) Data Type

Numeric data includes decimal numbers, whole numbers (stored as decimals), and special values.

## Regular Numbers

```{r}
# Regular numeric values
x <- 10.5
class(x)  # "numeric"
```

```{r}
y <- -3.7
class(y)  # "numeric"
```

## Special Numeric Values

**R** has two special numeric values:

- Inf - Positive infinity

- Inf - Negative infinity

- NaN - Not a Number (undefined mathematical result)

```{r}
# Infinity examples
1 / 0        # Returns Inf
class(Inf)   # "numeric"

-5 / 0       # Returns -Inf
class(-Inf)  # "numeric"
```

```{r}
# NaN (Not a Number) examples
0 / 0        # Returns NaN
class(NaN)   # "numeric"

Inf - Inf    # Also returns NaN
```

## Converting to Numeric

Use `as.numeric()` to convert other types to numeric:

```{r}
# Character to numeric
as.numeric("123")      # 123
class(as.numeric("123"))

# Logical to numeric (TRUE = 1, FALSE = 0)
as.numeric(TRUE)       # 1
as.numeric(FALSE)      # 0
```

# Integer Data Type

Integers are whole numbers without decimal points. They use less memory than numeric values.

## Creating Integers

Add the letter L after a number to create an integer:

```{r}
# Creating integers
age <- 25L
class(age)          # "integer"

# Using as.integer()
height <- as.integer(68.7)
height              # 68 (decimal portion dropped)
class(height)       # "integer"
```

## Integer vs. Numeric

```{r}
# Memory efficiency example
numeric_vec <- c(1, 2, 3)
integer_vec <- c(1L, 2L, 3L)
```

```{r}
object.size(numeric_vec)  # Larger memory usage
object.size(integer_vec)  # Smaller memory usage
```

**Note:** When working with large datasets (millions of rows), using integers for ID columns, counts, or indices saves memory and improves speed.

# Character (Text) Data Type

Character data stores text strings. Always enclose text in quotes.

## Creating Character Data

```{r}
# Single or double quotes both work
name <- "R programming"
city <- 'New York'
empty_string <- ""

class(name)         # "character"
class(city)         # "character"
class(empty_string) # "character"
```

## Converting to Character
```{r}
# Number to character
as.character(123)       # "123"
as.character(3.14)      # "3.14"

# Logical to character
as.character(TRUE)      # "TRUE"
as.character(FALSE)     # "FALSE"
```

## Practical Text Operations

```{r}
# Concatenating (combining) text
greeting <- "Hello"
name <- "Alice"
paste(greeting, name)   # "Hello Alice"

# Getting text length
nchar("R is great")      # 11 characters (including spaces)
```

# Logical (Boolean) Data Type

Logical values represent truth: either TRUE or FALSE.

## Creating Logical Values

```{r}
# Direct assignment
is_raining <- TRUE
has_umbrella <- FALSE

class(is_raining)   # "logical"
```

## Logical Comparisons

Common comparison operators:
Operator	Meaning	Example

| Operator | Meaning | Example |
|----------|---------|---------|
| `>` | Greater than | `5 > 3` → `TRUE` |
| `<` | Less than | `5 < 3` → `FALSE` |
| `>=` | Greater than or equal | `5 >= 5` → `TRUE` |
| `<=` | Less than or equal | `3 <= 2` → `FALSE` |
| `==` | Equal to | `10 == 10` → `TRUE` |
| `!=` | Not equal to | `10 != 5` → `TRUE` |

```{r}
# Comparison examples
10 > 5          # TRUE
10 > 10         # FALSE
10 >= 10        # TRUE
"apple" == "apple"   # TRUE
"apple" != "orange"  # TRUE
```

## Logical Operators

Combine logical values using:

    & = AND (both must be TRUE)

    | = OR (at least one must be TRUE)

    ! = NOT (reverses the value)

```{r}
# AND operator - both conditions must be TRUE
(5 > 3) & (2 < 4)     # TRUE (both conditions met)
(5 > 3) & (2 > 4)     # FALSE (second condition fails)

# OR operator - at least one condition must be TRUE
(5 > 3) | (2 > 4)     # TRUE (first condition is TRUE)
(5 < 3) | (2 > 4)     # FALSE (both conditions are FALSE)

# NOT operator - reverses TRUE/FALSE
!TRUE                 # FALSE
!FALSE                # TRUE
!(5 > 3)              # FALSE
```

## Conversion Rules

```{r}
# Numeric to logical: 0 = FALSE, everything else = TRUE
as.logical(0)        # FALSE
as.logical(1)        # TRUE
as.logical(-5)       # TRUE
as.logical(3.14)     # TRUE

# Logical to numeric: TRUE = 1, FALSE = 0
as.numeric(TRUE)     # 1
as.numeric(FALSE)    # 0
```

# Complex Data Type

Complex numbers have real and imaginary parts (rarely needed in basic data analysis).

```{r}
# Creating complex numbers
comp1 <- 2 + 3i
comp2 <- 4 - 2i

class(comp1)    # "complex"

# Square root of negative number requires complex
sqrt(-1)        # NaN
sqrt(-1 + 0i)   # 0+1i (complex result)
```

# Practical Examples Using class()
## Example 1: Checking Data Type After Operations

```{r}
# What type does a mathematical operation produce?
result1 <- 10 / 3
class(result1)  # "numeric"

result2 <- 10L + 5L
class(result2)  # "integer" (adding integers keeps integer type)

result3 <- 10L + 5.5
class(result3)  # "numeric" (integer + numeric = numeric)
```

## Example 2: Type Conversion Chain

```{r}
# Starting with character
text_number <- "42"
class(text_number)  # "character"

# Convert to integer
int_number <- as.integer(text_number)
class(int_number)   # "integer"

# Convert to numeric
num_number <- as.numeric(int_number)
class(num_number)   # "numeric"

# Convert to logical
log_number <- as.logical(num_number)
class(log_number)   # "logical"
```

## Example 3: Detecting Data Type Problems

```{r}
# Common real-world issue: numbers stored as text
sales_data <- c("100", "200", "300", "400")
class(sales_data)  # "character" - problem: can't do math!

# Solution: convert to numeric
sales_numeric <- as.numeric(sales_data)
class(sales_numeric)  # "numeric" - now we can calculate
mean(sales_numeric)   # 250
```

## Example 4: Class of Mixed Operations

```{r}
# When mixing types, R follows promotion rules
mixed1 <- c(1, 2, "three")
class(mixed1)  # "character" (character dominates)

mixed2 <- c(TRUE, FALSE, 1)
class(mixed2)  # "numeric" (coerced to numbers: TRUE=1, FALSE=0)

mixed3 <- c(TRUE, "FALSE")
class(mixed3)  # "character" (character dominates)
```

# Summary Table

| Type | `class()` | `typeof()` | Example | Common Uses |
|------|-----------|------------|---------|--------------|
| Numeric | "numeric" | "double" | `3.14` | Measurements, decimals |
| Integer | "integer" | "integer" | `42L` | Counts, indices, IDs |
| Character | "character" | "character" | `"hello"` | Text, categories |
| Logical | "logical" | "logical" | `TRUE` | Conditions, filters |
| Complex | "complex" | "complex" | `2+3i` | Engineering, math |

# Exercises

## Exercise 1: Identify Data Types

What will class() return for each of these values? Write your answer, then uncomment and run the code to check.

```{r}
# a) class(42)
# Your answer: ___________

# b) class("42")
# Your answer: ___________

# c) class(42L)
# Your answer: ___________
```

### Solution 

```{r}
class(42)        # "numeric"
class("42")      # "character"
class(42L)       # "integer"
```

## Exercise 2: Type Conversion Practice

Complete the code below to convert the character "3.14" to numeric, then check its class.

```{r}
# Convert "3.14" to numeric and store in variable 'pi_value'
# Then check its class
# (Write your code here, then uncomment to run)
# pi_value <- 
# class(pi_value)
```

## Solution 

```{r}
pi_value <- as.numeric("3.14")
class(pi_value)  # "numeric"
```

## Exercise 3: Logical Operations

Predict the class and value of the following expressions. Check your answers by running the code.

```{r}
# a) class(5 > 3)
# Prediction: ___________

# b) class(as.numeric(TRUE))
# Prediction: ___________

# c) class(1 == "1")
# Prediction: ___________
```

## Solution 

```{r}
# a) "logical" (result is TRUE)
# b) "numeric" (TRUE converts to 1)
# c) "logical" (R compares after converting "1" to 1)
```

## Exercise 4: Real-world Scenario

A dataset has a column of student grades as characters: c("85", "92", "78", "fail", "96"). You need to calculate the average grade. What happens if you try mean(grades) directly? How would you fix it?

```{r}
grades <- c("85", "92", "78", "fail", "96")

# Try calculating mean (uncomment to see what happens)
# mean(grades)

# Write code to convert valid grades to numeric and handle the "fail" value
# Then calculate the average of valid grades
```

## Solution 

```{r}
grades <- c("85", "92", "78", "fail", "96")
grades_numeric <- as.numeric(grades)  # "fail" becomes NA
mean(grades_numeric, na.rm = TRUE)    # 87.75 (average of 85, 92, 78, 96)
```

## Exercise 5: Create Your Own Examples

Create one example for each data type using the <- assignment operator. Then use class() to verify the type.

```{r}
# Numeric example
# Your code here

# Integer example
# Your code here

# Character example
# Your code here

# Logical example
# Your code here
```

## Solution 

```{r}
my_numeric <- 3.14159
class(my_numeric)  # "numeric"

my_integer <- 100L
class(my_integer)  # "integer"

my_text <- "Data Science"
class(my_text)     # "character"

my_logic <- 10 > 5
class(my_logic)    # "logical"
```

## Exercise 6: Type Hierarchy Challenge

When combining different data types in a vector, **R** follows a hierarchy: logical → integer → numeric → character. Test this by creating vectors with mixed types and checking their class.

```{r}
# Example: combine logical and numeric
mix1 <- c(TRUE, 5)
class(mix1)  # Run this - what type do you get?

# Now try: combine integer and character
mix2 <- c(10L, "test")
class(mix2)  # What type?

# Now try: combine logical and character
mix3 <- c(FALSE, "R")
class(mix3)  # What type?

# Write your observations here as comments
```

## Solution 

```{r}
mix1 <- c(TRUE, 5)        # "numeric" (logical promoted to numeric)
mix2 <- c(10L, "test")    # "character" (character dominates)
mix3 <- c(FALSE, "R")     # "character" (character dominates)
```

