---
title: '__Topic 5: Arrays and Matrices in R__'
author: "Esam Mahdi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to Arrays and Matrices

In the previous lecture, we learned about **vectors** - one-dimensional collections of data. But real-world data often has **multiple dimensions**: sales figures by month and product category, temperature readings by city and day, or test scores by student and subject.

**Arrays** extend vectors to multiple dimensions. A **matrix** is a special case of an array with exactly **two dimensions** (rows and columns).

Think of dimensions this way:

  - **Vector:** A single row of boxes (1D)

  - **Matrix:** A grid of boxes with rows and columns (2D)

  - **Array:** A cube or hypercube of boxes (3D or more)

```
Vector (1D):     [10] [20] [30] [40] [50]

Matrix (2D):      [10] [20] [30]
                  [40] [50] [60]
                  [70] [80] [90]

Array (3D):      Two or more matrices stacked like a book
```

## Creating Arrays

**The `array()` Function**

The most flexible way to create an array is with the `array()` function:

```{r}
# Create a 3D array: 3 rows, 4 columns, 2 layers
my_array <- array(1:24, dim = c(3, 4, 2))
my_array
```

**Understanding the output:**

  - , , 1 means "first layer" (first matrix)

  - , , 2 means "second layer" (second matrix)

  - The numbers show: [row, column, layer]

## Creating a Matrix with `matrix()`

For 2D data, `matrix()` is more convenient:

```{r}
# Create a 3x4 matrix filled with numbers 1 through 12
my_matrix <- matrix(1:12, nrow = 3, ncol = 4)
my_matrix
```

## The Dimension Attribute

You can also convert a vector into an array by giving it a `dim` attribute:

```{r}
# Start with a vector
z <- 1:24
z
# Give it dimensions (3 rows, 4 columns, 2 layers)
dim(z) <- c(3, 4, 2)
z  # Now it's a 3D array!
```

## The Recycling Rule

If your data vector is shorter than the array size, **R** recycles (repeats) the values:

```{r}
# Only provide 6 numbers, but need 12 for a 3x4 matrix
short_data <- c(1, 2, 3, 4, 5, 6)
recycled_matrix <- matrix(short_data, nrow = 3, ncol = 4)
recycled_matrix  # R repeats 1,2,3,4,5,6 to fill all 12 positions

# Create an array of all zeros (very common!)
zero_array <- array(0, dim = c(3, 4, 2))
zero_array
```

## Column-Major Order (Important!)

**R** fills arrays in column-major order (like FORTRAN, not like C/Python). This means the **first subscript moves fastest**:

```{r}
# A 3x4 matrix filled with 1:12
example <- matrix(1:12, nrow = 3, ncol = 4)
example

# The order of filling is:
# [1,1]=1, [2,1]=2, [3,1]=3,  (first column)
# [1,2]=4, [2,2]=5, [3,2]=6,  (second column)
# [1,3]=7, [2,3]=8, [3,3]=9,  (third column)
# [1,4]=10, [2,4]=11, [3,4]=12 (fourth column)
```

To fill by rows instead (like a spreadsheet), use `byrow = TRUE`:

```{r}
row_filled <- matrix(1:12, nrow = 3, ncol = 4, byrow = TRUE)
row_filled
```

## Matrix Indexing

Matrix indexing allows you to extract specific rows, columns, or individual values from a matrix. The basic syntax is `matrix[row_index, column_index]`.

```{r}
# Create a sample matrix
M <- matrix(1:20, nrow = 4, ncol = 5, byrow = TRUE)
rownames(M) <- c("Row1", "Row2", "Row3", "Row4")
colnames(M) <- c("Col1", "Col2", "Col3", "Col4", "Col5")
M

# Extract a single value (row 2, column 3)
M[2, 3]

# Extract an entire row (all columns of row 3)
M[3, ]

# Extract an entire column (all rows of column 4)
M[, 4]

# Extract multiple specific rows and columns
M[c(1, 3), c(2, 4)]

# Extract a range of rows and columns
M[2:4, 3:5]

# Exclude rows or columns (negative indexing)
M[-1, ]        # All rows except row 1
M[, -c(2, 4)]  # All columns except columns 2 and 4
M[-2, -3]      # All rows except row 2, all columns except column 3
```

**Understanding Different Indexing Patterns**

```
# x[,3] means extract all values in column 3
M[, 3]

# x[-1, c(2,7)] means extract all rows except the 1st one, 
# and only columns 2 and 7 (if column 7 exists)
M[-1, c(2, 5)]  # Column 5 instead of 7 (since we only have 5 columns)

# Using logical indexing
M[M > 15]  # Returns all values greater than 15 as a vector

# Using column names
M[, "Col3"]
M["Row2", ]
M["Row2", "Col4"]
```

## Accessing Array Elements: Indexing

**Single Element Access**

Use square brackets with commas separating dimensions:

```{r}
# Create a 4x5 matrix
sales <- matrix(1:20, nrow = 4, ncol = 5)
sales

# Get element at row 2, column 3
sales[2, 3]

# Get element at row 4, column 5
sales[4, 5]
```

## Entire Rows or Columns

Leave a dimension blank to get all elements in that dimension:

```{r}
# Get entire 2nd row (all columns)
sales[2, ]

# Get entire 3rd column (all rows)
sales[, 3]

# Get entire matrix (all rows, all columns)
sales[, ]
```

## Ranges and Submatrices

Use `:` to specify ranges:

```{r}
# Get rows 1-3 and columns 2-4
sales[1:3, 2:4]

# Get rows 2-4 and column 5 only
sales[2:4, 5]

# Get rows 1 and 3, and columns 2, 4, and 5
sales[c(1, 3), c(2, 4, 5)]
```

## Excluding Elements

Use `-` to exclude rows or columns:

```{r}
# Exclude row 2
sales[-2, ]

# Exclude columns 3 and 4
sales[, -c(3, 4)]

# Exclude row 1 and column 5
sales[-1, -5]
```

## Advanced Indexing with Matrices

**Using an Index Matrix** 

Sometimes you need to extract or modify specific elements that don't form a neat rectangle. Use an **index matrix** - a matrix where each row contains the coordinates of one element.

```{r}
# Create a 4x5 array
X <- array(1:20, dim = c(4, 5))
X

# We want to extract and modify:
# X[1,3], X[2,2], X[3,1]

# Create an index matrix (3 rows, 2 columns)
i <- matrix(c(1, 3,   # Row 1, Column 3
              2, 2,   # Row 2, Column 2
              3, 1),  # Row 3, Column 1
            nrow = 3, ncol = 2, byrow = TRUE)
i

# Extract the three elements
X[i]

# Replace those three elements with zeros
X[i] <- 0
X
```

**Practical Example: Block Design Matrix**

```{r}
# Suppose we have 10 plots, 3 blocks, and 4 varieties
n <- 10  # plots
b <- 3   # blocks
v <- 4   # varieties

# Create design matrices (initialize with zeros)
Xb <- matrix(0, n, b)  # Block design matrix
Xv <- matrix(0, n, v)  # Variety design matrix

# Suppose these are our assignments
blocks <- c(1, 2, 1, 3, 2, 1, 3, 2, 3, 1)
varieties <- c(2, 1, 3, 2, 4, 1, 3, 4, 2, 1)

# Create index matrices
ib <- cbind(1:n, blocks)  # Plot number and block assignment
iv <- cbind(1:n, varieties)  # Plot number and variety assignment

# Fill the design matrices
Xb[ib] <- 1
Xv[iv] <- 1

# View results
head(Xb)
head(Xv)

# Create incidence matrix (which varieties appear in which blocks)
N <- t(Xb) %*% Xv  # Matrix multiplication
N  # Shows counts of each variety in each block
```

# Array Operations

## Element-wise Operations

Arrays support the same vectorized operations as vectors:

```{r}
A <- matrix(1:9, nrow = 3, ncol = 3)
B <- matrix(10:18, nrow = 3, ncol = 3)

# Element-wise addition
A + B

# Element-wise multiplication (NOT matrix multiplication!)
A * B

# Element-wise division
B / A

# Scalar operations apply to all elements
A * 2
A + 100
A ^ 2
```

## Matrix Multiplication (`%*%`)

For true matrix multiplication, use `%*%`:

```{r}
C <- matrix(1:6, nrow = 2, ncol = 3)  # 2x3 matrix
D <- matrix(7:12, nrow = 3, ncol = 2) # 3x2 matrix

C
D

# Matrix multiplication: (2x3) %*% (3x2) = (2x2)
C %*% D

# Matrix multiplication is NOT commutative
# D %*% C  # This would give an error (dimensions don't match)
```

# Matrix-Specific Functions

## Basic Information

```{r}
M <- matrix(1:12, nrow = 3, ncol = 4)

# Dimensions
nrow(M)  # Number of rows
ncol(M)  # Number of columns
dim(M)   # Both dimensions

# Check if it's a matrix
is.matrix(M)

# Get row and column names (initially NULL)
rownames(M)
colnames(M)

# Add names
rownames(M) <- c("Row1", "Row2", "Row3")
colnames(M) <- c("Col1", "Col2", "Col3", "Col4")
M
```

## Summary Statistics on Matrices

```{r}
M <- matrix(1:12, nrow = 3, ncol = 4)
M

# Minimum and maximum of entire matrix
min(M)
max(M)

# Sum of all elements
sum(M)

# Row-wise operations
rowSums(M)    # Sum of each row
rowMeans(M)   # Mean of each row

# Column-wise operations
colSums(M)    # Sum of each column
colMeans(M)   # Mean of each column
```

**The `apply()` Function**

The `apply()` function is a powerful tool for applying a function to margins (rows or columns) of a matrix or array. The syntax is:

`apply(X, MARGIN, FUN, ...)`

  - X: your matrix or array

  - MARGIN: 1 for rows, 2 for columns, c(1,2) for both

  - FUN: the function to apply (e.g., sum, mean, sd, etc.)

```{r}
M <- matrix(1:12, nrow = 3, ncol = 4)
M

# Apply sum to rows (MARGIN = 1)
apply(M, 1, sum)  # Same as rowSums(M)

# Apply mean to columns (MARGIN = 2)
apply(M, 2, mean)  # Same as colMeans(M)

# Apply custom function to rows
apply(M, 1, function(x) max(x) - min(x))

# Apply to both dimensions (returns original matrix)
apply(M, c(1,2), function(x) x^2)
```

**Transpose (`t()`)**

```{r}
M <- matrix(1:6, nrow = 2, ncol = 3)
M

# Transpose (swap rows and columns)
t(M)
```

**Diagonal (`diag()`)**

The `diag()` function behaves differently depending on its input:

```{r}
# If given a vector, creates a diagonal matrix
diag(c(10, 20, 30))

# If given a matrix, extracts the diagonal
M <- matrix(1:9, nrow = 3)
M
diag(M)  # Returns c(1, 5, 9)

# If given a single number, creates an identity matrix
diag(4)  # 4x4 identity matrix
```

**Trace (Sum of Diagonal Elements)**

**R** doesn't have a built-in trace function, but we can create one:

```{r}
# Create a square matrix
M <- matrix(1:9, nrow = 3)
M

# Calculate the trace (sum of diagonal elements)
sum(diag(M))  # 1 + 5 + 9 = 15

# Another example
M2 <- matrix(c(4, 2, 1, 3, 7, 2, 5, 1, 6), nrow = 3)
M2
sum(diag(M2))  # 4 + 7 + 6 = 17
```

## Advanced Matrix Operations

**Solving Linear Equations (`solve()`)**

Solve the system: `A %*% x = b`

```{r}
# Create a 3x3 matrix A
A <- matrix(c(3, 1, 1,
              1, 2, 1,
              1, 1, 2), nrow = 3, byrow = TRUE)
A

# Create a vector b
b <- c(5, 5, 6)

# Solve for x
x <- solve(A, b)
x

# Verify: A %*% x should equal b
A %*% x
```

## Matrix Inverse

```{r}
# Compute the inverse of A
A_inv <- solve(A)
A_inv

# Verify: A %*% A_inv = Identity matrix
round(A %*% A_inv, 2)  # Should be identity matrix
```

Important Note: It's more efficient and numerically stable to use `solve(A, b)` rather than `solve(A) %*% b`.

## Eigenvalues and Eigenvectors

**What are Eigenvalues and Eigenvectors?**

In linear algebra, eigenvectors are special vectors that only get scaled (not rotated) when a linear transformation is applied to them. The eigenvalue is the scaling factor.

  - **Eigenvector:** A direction that remains unchanged under a transformation

  - **Eigenvalue:** The factor by which the eigenvector is stretched or compressed

Applications include:

  - Principal Component Analysis (PCA) for dimensionality reduction

  - Google's PageRank algorithm

    Image compression

  - Stability analysis in engineering
    
```{r}
# Create a symmetric matrix
S <- matrix(c(4, 2, 1,
              2, 3, 2,
              1, 2, 4), nrow = 3, byrow = TRUE)
S

# Compute eigenvalues and eigenvectors
eigen_result <- eigen(S)

# Eigenvalues
eigen_result$values

# Eigenvectors (columns of this matrix)
eigen_result$vectors

# Verify: For first eigenvector, S %*% v = λ * v
lambda_1 <- eigen_result$values[1]
v1 <- eigen_result$vectors[, 1]
S %*% v1  # This should equal
lambda_1 * v1  # This

# Get only eigenvalues (more efficient)
eigen(S, only.values = TRUE)$values
```

## Cross Products

`crossprod()` is more efficient than `t(X) %*% Y`:

```{r}
X <- matrix(runif(20), nrow = 5, ncol = 4)
Y <- matrix(runif(15), nrow = 5, ncol = 3)

# These are equivalent, but crossprod is faster
identical(t(X) %*% Y, crossprod(X, Y))

# Cross product of X with itself
crossprod(X)  # Same as t(X) %*% X
```

## Permuting Array Dimensions (`aperm()`)

The `aperm()` function rearranges the dimensions of an array (like transposing for higher dimensions):

```{r}
# Create a 3D array (2x3x4)
original <- array(1:24, dim = c(2, 3, 4))
original

# Permute dimensions: new order is (3, 1, 2)
# So original dimension 3 becomes new dimension 1
# original dimension 1 becomes new dimension 2
# original dimension 2 becomes new dimension 3
permuted <- aperm(original, perm = c(3, 1, 2))
dim(original)   # 2 3 4
dim(permuted)   # 4 2 3

# For a matrix, aperm() with c(2,1) is transpose
M <- matrix(1:6, nrow = 2, ncol = 3)
M
aperm(M, c(2, 1))  # Same as t(M)
```

## Combining and Converting Matrices

**`rbind()` and `cbind()` $-$ Combining Matrices**

The `rbind()` function combines matrices by **binding rows** (stacking vertically).
The `cbind()` function combines matrices by **binding columns** (joining horizontally).

```{r}
# Create two matrices
matrix1 <- matrix(1:6, nrow = 2, ncol = 3)
matrix2 <- matrix(7:12, nrow = 2, ncol = 3)

matrix1
matrix2

# Stack vertically (row bind)
rbind(matrix1, matrix2)

# Join horizontally (column bind)
cbind(matrix1, matrix2)

# Combine different dimensions
matrix3 <- matrix(13:18, nrow = 3, ncol = 2)
matrix1
matrix3
```

```
# Row bind with different row counts (number of columns of matrices must match)

rbind(matrix1, matrix3)  # matrix1 has 3 columns, matrix3 has 2 columns? This will give an error!
```

```{r}
# Make them compatible first
matrix3_compatible <- matrix(13:18, nrow = 3, ncol = 3)
rbind(matrix1, matrix3_compatible)

# Adding a vector as a new row
new_row <- c(100, 110, 120)
rbind(matrix1, new_row)

# Adding a vector as a new column
new_col <- c(200, 210)
cbind(matrix1, new_col)
```

**Converting Matrices to Vectors with `as.vector()`**

The `as.vector()` function converts a matrix into a vector **by columns** (column-major order):

```{r}
# Create a matrix
M <- matrix(1:6, nrow = 2, ncol = 3)
M

# Convert to vector (column by column)
as.vector(M)  # Returns: 1,2,3,4,5,6 (down columns)

# Compare with filling by rows
M_row <- matrix(1:6, nrow = 2, ncol = 3, byrow = TRUE)
M_row
as.vector(M_row)  # Still goes down columns: 1,4,2,5,3,6

# To get row-wise conversion, transpose first
as.vector(t(M))  # Row-wise: 1,3,5,2,4,6? No, let's see
t(M)  # Transpose first
as.vector(t(M))  # Now row-wise of original
```

**Lists vs. Arrays/Matrices**

A **list** is a different type of data structure that can hold elements of **different types** (unlike vectors, matrices, and arrays which require all elements to be the same type).

**Key differences:**

  - **Matrix/Array:** All elements must be the same type (homogeneous)

  - **List:** Can contain different types (heterogeneous) - numbers, text, matrices, even other lists
  
```{r}
# Matrix requires same type (all numeric)
matrix(1:5, nrow = 1)  # Works fine

# But mixing types causes coercion
mixed_matrix <- matrix(c(1, "two", 3, "four"), nrow = 2)
mixed_matrix  # Numbers become text

# Lists can handle different types naturally
my_list <- list(
  name = "John",
  age = 25,
  scores = c(85, 90, 88),
  matrix = matrix(1:4, nrow = 2)
)
my_list

# Access list elements with [[ ]]
my_list[[1]]  # First element
my_list$name  # Access by name
my_list$scores

# You can store matrices inside lists
list_of_matrices <- list(
  M1 = matrix(1:4, nrow = 2),
  M2 = matrix(5:8, nrow = 2)
)
list_of_matrices
list_of_matrices$M1
```
   
**Example: Sales Data Analysis**

```{r}
# Sales data: 4 quarters, 3 products, 2 regions
sales <- array(c(120, 135, 110, 140,   # Product 1, Region 1
                 95, 105, 115, 100,    # Product 2, Region 1
                 80, 90, 85, 95,       # Product 3, Region 1
                 100, 115, 105, 120,   # Product 1, Region 2
                 85, 95, 90, 100,      # Product 2, Region 2
                 70, 80, 75, 85),      # Product 3, Region 2
               dim = c(4, 3, 2))
dimnames(sales) <- list(
  Quarter = c("Q1", "Q2", "Q3", "Q4"),
  Product = c("A", "B", "C"),
  Region = c("North", "South")
)
sales

# Total sales by product (using apply)
apply(sales, 2, sum)  # Sum over quarters and regions

# Total sales by region (using apply)
apply(sales, 3, sum)

# Quarterly sales for Product A in North region
sales[, "A", "North"]

# Best quarter for each product-region combination
apply(sales, c(2, 3), which.max)
```

# Common Mistakes

## Mistake 1: Confusing `*` and `%*%`

```{r}
A <- matrix(1:4, nrow = 2)
B <- matrix(5:8, nrow = 2)

# Element-wise multiplication (probably not what you want for linear algebra)
A * B

# Matrix multiplication (what you usually want for linear algebra)
A %*% B
```

## Mistake 2: Dimension Mismatch

```{r}
A <- matrix(1:6, nrow = 2, ncol = 3)
B <- matrix(7:12, nrow = 2, ncol = 3)

# This works (element-wise)
A * B

# This fails (matrix multiplication requires inner dimensions to match)
# A %*% B  # Error: non-conformable arguments

# Correct:
C <- matrix(7:12, nrow = 3, ncol = 2)  # 3x2 matrix
A %*% C  # 2x3 %*% 3x2 = 2x2 (works!)
```

## Mistake 3: Forgetting Column-Major Order

```{r}
# You might expect this to be:
# [1,1]=1, [1,2]=2, [1,3]=3, [2,1]=4, etc.
expected <- matrix(1:6, nrow = 2, ncol = 3, byrow = TRUE)

# But without byrow, you get column-major:
actual <- matrix(1:6, nrow = 2, ncol = 3)

cat("With byrow=TRUE (like spreadsheet):\n")
print(expected)
cat("\nWithout byrow (column-major):\n")
print(actual)
```

## Mistake 4: Off-by-One in Indexing

```
M <- matrix(1:9, nrow = 3)

# Wrong (indexing starts at 1, not 0)
# M[0, 1]  # Returns empty matrix, not row 1

# Correct
M[1, 1]  # First element

# Wrong (trying to access beyond dimensions)
# M[5, 3]  # Error: subscript out of bounds
```

# Summary Table

| Operation | Syntax | Example | Purpose |
|-----------|--------|---------|---------|
| Create array | `array()` | `array(1:24, dim = c(3, 4, 2))` | Multi-dimensional array |
| Create matrix | `matrix()` | `matrix(1:12, nrow = 3, ncol = 4)` | 2D array |
| Set dimensions | `dim()` | `dim(z) <- c(3, 4, 2)` | Convert vector to array |
| Access element | `[i, j]` | `A[2, 3]` | Single element |
| Entire row | `[i, ]` | `A[2, ]` | All columns of row i |
| Entire column | `[, j]` | `A[, 3]` | All rows of column j |
| Submatrix | `[i:j, k:l]` | `A[1:3, 2:4]` | Range of rows and columns |
| Exclude | `[-i, -j]` | `A[-1, -2]` | Remove rows/columns |
| Index matrix | `matrix[c(...)]` | `X[index_matrix]` | Irregular extraction |
| Element-wise mult | `*` | `A * B` | Multiply corresponding elements |
| Matrix multiplication | `%*%` | `A %*% B` | Linear algebra multiplication |
| Outer product | `%o%` | `x %o% y` | All combinations |
| Transpose | `t()` | `t(A)` | Swap rows and columns |
| Diagonal | `diag()` | `diag(A)` | Extract/set diagonal |
| Solve equations | `solve()` | `solve(A, b)` | Solve Ax = b |
| Inverse | `solve()` | `solve(A)` | Compute A⁻¹ |
| Eigenvalues | `eigen()` | `eigen(A)` | Eigen decomposition |
| Permute dimensions | `aperm()` | `aperm(A, c(2, 1))` | Rearrange dimensions |
| Cross product | `crossprod()` | `crossprod(X, Y)` | Efficient X'Y |
| Bind rows | `rbind()` | `rbind(M1, M2)` | Stack matrices vertically |
| Bind columns | `cbind()` | `cbind(M1, M2)` | Join matrices horizontally |
| Matrix to vector | `as.vector()` | `as.vector(M)` | Convert to vector (by column) |
| Create list | `list()` | `list(1, "a", M)` | Heterogeneous collection |

# Exercises

## Exercise 1: Create and Explore a Matrix

Create a 4x6 matrix filled with numbers 1 through 24 (row by row). Then:

  - Extract the element at row 3, column 5

  - Extract the entire 2nd row

  - Extract the entire 4th column

  - Extract a submatrix of rows 2-3 and columns 3-5

```{r}
# Your code here:
```

## Solution

```{r}
# Create matrix by row
M <- matrix(1:24, nrow = 4, ncol = 6, byrow = TRUE)
M

# Extract element
M[3, 5]  # Row 3, Column 5

# Entire 2nd row
M[2, ]

# Entire 4th column
M[, 4]

# Submatrix rows 2-3, columns 3-5
M[2:3, 3:5]
```

## Exercise 2: Matrix Operations

Given the following matrices:

```{r}
A <- matrix(c(2, 1, 0, -1, 3, 2, 4, -2, 1), nrow = 3, byrow = TRUE)
B <- matrix(c(1, 2, 1, 0, 1, -1, 3, 2, 0), nrow = 3, byrow = TRUE)
```

Calculate:

  - A + B

  - A %*% B

  - The transpose of A

  - The inverse of A (if it exists)

  - The trace of A

  - Row sums and column sums of A

```{r}
# Your code here:
```

## Solution 

```{r}
A <- matrix(c(2, 1, 0, -1, 3, 2, 4, -2, 1), nrow = 3, byrow = TRUE)
B <- matrix(c(1, 2, 1, 0, 1, -1, 3, 2, 0), nrow = 3, byrow = TRUE)

# A + B
A + B

# A %*% B
A %*% B

# Transpose of A
t(A)

# Inverse of A (check determinant first)
det(A)  # Non-zero means invertible
solve(A)

# Trace of A
sum(diag(A))

# Row sums and column sums
rowSums(A)
colSums(A)
```

## Exercise 3: Solving Linear Equations

Solve the following system of equations using **R**:
text

```
2x + y + z = 7
x - y + 2z = 4
3x + 2y - z = 5
```

```{r}
# Your code here:
```

## Solution 

```{r}
# Coefficient matrix
A <- matrix(c(2, 1, 1,
              1, -1, 2,
              3, 2, -1), nrow = 3, byrow = TRUE)

# Right-hand side
b <- c(7, 4, 5)

# Solve
solve(A, b)  # x = 2, y = 1, z = 1
```

## Exercise 4: Working with a 3D Array

Create a 3D array with dimensions $(4, 5, 3)$ containing random numbers between 0 and 100. Then:

  - Calculate the mean of each layer (3rd dimension) using apply()

  - Calculate the same means without apply() (hint: use colMeans or another approach)

  - Find the maximum value in each row (across all columns and layers)

  - Calculate the column sums for the second layer only
  
```{r}
# Your code here:
```

## Solution 

```{r}
set.seed(123)
arr <- array(runif(4*5*3, 0, 100), dim = c(4, 5, 3))

# Mean of each layer using apply()
apply(arr, 3, mean)

# Mean of each layer without apply() - approach 1
mean_layer1 <- mean(arr[, , 1])
mean_layer2 <- mean(arr[, , 2])
mean_layer3 <- mean(arr[, , 3])
c(mean_layer1, mean_layer2, mean_layer3)

# Mean of each layer without apply() - approach 2 using colMeans then rowMeans
# First average across columns, then across rows within each layer
sapply(1:3, function(i) mean(colMeans(arr[, , i])))

# Maximum in each row (across all columns and layers)
apply(arr, 1, max)

# Column sums for the second layer only
colSums(arr[, , 2])
```

## Exercise 5: Using `rbind()` and `cbind()`

Create three vectors: `names = c("A", "B", "C", "D"), scores1 = c(85, 90, 78, 92), scores2 = c(88, 87, 81, 95)`. Then:

  - Combine names and scores1 into a 4x2 matrix using `cbind()`

  - Combine scores1 and scores2 into an 8x1 matrix using `rbind()`

  - Add a new row with values `c("E", 89, 94)` to your first matrix

  - Convert your combined matrix to a vector using `as.vector()`

```{r}
# Your code here:
```

## Solution

```{r}
# Create vectors
names <- c("A", "B", "C", "D")
scores1 <- c(85, 90, 78, 92)
scores2 <- c(88, 87, 81, 95)

# Combine names and scores1 into a matrix
matrix1 <- cbind(names, scores1)
matrix1

# Combine scores1 and scores2 using rbind
matrix2 <- rbind(scores1, scores2)
matrix2

# Add a new row to matrix1
new_row <- c("E", 89, 94)
matrix1 <- rbind(matrix1, new_row)
matrix1

# Convert to vector
as.vector(matrix1)
```

## Exercise 6: Matrix Indexing Practice

Create a 5x5 matrix with numbers 1 through 25 (row by row). Then extract:

  - All values in column 3

  - All rows except the first, but only columns 2 and 4

  - The 2x2 submatrix at the bottom-right corner

  - All values greater than 15 as a vector

```{r}
# Your code here:
```

## Solution

```{r}
# Create 5x5 matrix with numbers 1-25 (row by row)
M <- matrix(1:25, nrow = 5, ncol = 5, byrow = TRUE)
M

# All values in column 3
M[, 3]
```

